---
title: "LOO-PSIS: Model Comparison with Cross-Validation"
subtitle: "Bayesian Mixed Effects Models with brms for Linguists"
date: today
author: "Workshop Materials"
format: 
  html:
    toc: true
    number-sections: true
    code-fold: false
    keep-md: true
    theme:
      light: flatly
      dark: darkly
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    code-fold: false
    keep-md: false
execute:
  cache: true
  warning: false
  message: false
---

# LOO-PSIS: Leave-One-Out Cross-Validation

LOO-PSIS (Leave-One-Out Cross-Validation with Pareto-Smoothed Importance Sampling) helps answer: **Which model predicts new data better?**

## Why Use LOO Instead of Prior Comparison?

These approaches answer different questions:

**Prior comparison** (what we did earlier):

- Shows if posterior coefficient estimates and effect sizes are sensitive to prior choice
- Good for: reporting robustness of conclusions
- Question: "Do my results depend on my priors?"

**LOO comparison** (this approach):

- Shows which model predicts better
- Good for: feature selection, model building
- Question: "Which model structure produces better predictions?"
- Can compare:
  - Different priors (e.g., narrow/regularizing vs. wide priors)
  - Different likelihoods (e.g., normal vs. lognormal)
  - Different model structures (e.g., with/without random slopes)

**You can do both:**

1. First: Compare different priors within same model structure (sensitivity analysis)
2. Then: Use LOO to compare different model structures with best priors (model selection)

## Why Use LOO Instead of Bayes Factors?

**LOO advantages:**

- Priors less important because we evaluate predictive performance on new data
- Number of samples less important - most uncertainty comes from the data itself
- More stable and interpretable

**Bayes factors:**

- Very sensitive to prior choice
- Sensitive to number of samples
- Harder to interpret (what does BF = 3.2 mean?)

## Setup

```{r}
#| label: setup
#| message: false

# Configure backend BEFORE loading brms
if (requireNamespace("cmdstanr", quietly = TRUE)) {
  tryCatch({
    cmdstanr::cmdstan_path()
    options(brms.backend = "cmdstanr")
  }, error = function(e) {
    options(brms.backend = "rstan")
  })
}

library(brms)
library(tidyverse)
library(bayesplot)
library(posterior)
library(loo)
library(patchwork)

# Set seed for reproducibility
set.seed(42)
```

## Create Four Test Datasets

We'll create four datasets to test how LOO behaves under different conditions:

```{r create-data}
#| cache: false

# Ensure tidyverse is loaded for pipe operator
library(dplyr)

# Helper function to generate RT data
generate_rt_data <- function(n_obs, with_random_effects = TRUE, seed = 42) {
  set.seed(seed)
  
  # Determine number of subjects and items (ensure at least 2 for factors)
  n_subj <- max(5, min(10, ceiling(n_obs / 8)))
  n_items <- max(5, min(10, ceiling(n_obs / 8)))
  
  # Create base structure ensuring balanced conditions
  n_per_cond <- ceiling(n_obs / 2)
  
  data <- dplyr::bind_rows(
    expand.grid(
      subject = 1:n_subj,
      item = 1:n_items,
      condition = "A"
    ) %>% dplyr::slice(1:n_per_cond),
    expand.grid(
      subject = 1:n_subj,
      item = 1:n_items,
      condition = "B"
    ) %>% dplyr::slice(1:(n_obs - n_per_cond))
  ) %>%
    dplyr::mutate(
      subject = factor(subject),
      item = factor(item),
      condition = factor(condition)
    )
  
  if (with_random_effects) {
    # Generate data WITH true random effects (STRONGER effects for clear demonstration)
    subj_intercept <- rnorm(n_subj, 0, 0.4)  # Increased from 0.2
    subj_slope <- rnorm(n_subj, 0, 0.2)       # Increased from 0.1
    item_intercept <- rnorm(n_items, 0, 0.3)  # Increased from 0.15
    
    data <- data %>%
      dplyr::mutate(
        re_subject_int = subj_intercept[as.numeric(subject)],
        re_subject_slope = subj_slope[as.numeric(subject)] * (condition == "B"),
        re_item = item_intercept[as.numeric(item)],
        log_rt = 6 + 0.15 * (condition == "B") + 
                 re_subject_int + re_subject_slope + re_item +
                 rnorm(n(), 0, 0.15),  # Reduced residual noise from 0.2
        rt = exp(log_rt)
      ) %>%
      dplyr::select(subject, item, condition, log_rt, rt)
  } else {
    # Generate data WITHOUT random effects (pure fixed effect + noise)
    data <- data %>%
      dplyr::mutate(
        log_rt = 6 + 0.15 * (condition == "B") + rnorm(n(), 0, 0.4),  # Increased noise
        rt = exp(log_rt)
      )
  }
  
  return(data)
}

# Generate four datasets
rt_data_100_with <- generate_rt_data(100, with_random_effects = TRUE, seed = 123)
rt_data_100_without <- generate_rt_data(100, with_random_effects = FALSE, seed = 124)
rt_data_40_with <- generate_rt_data(40, with_random_effects = TRUE, seed = 125)
rt_data_40_without <- generate_rt_data(40, with_random_effects = FALSE, seed = 126)

# Create summary table
data_summaries <- data.frame(
  Scenario = c("n=100, WITH RE", "n=100, WITHOUT RE", "n=40, WITH RE", "n=40, WITHOUT RE"),
  N = c(nrow(rt_data_100_with), nrow(rt_data_100_without), 
        nrow(rt_data_40_with), nrow(rt_data_40_without)),
  Mean_logRT = c(
    round(mean(rt_data_100_with$log_rt), 2),
    round(mean(rt_data_100_without$log_rt), 2),
    round(mean(rt_data_40_with$log_rt), 2),
    round(mean(rt_data_40_without$log_rt), 2)
  ),
  SD_logRT = c(
    round(sd(rt_data_100_with$log_rt), 3),
    round(sd(rt_data_100_without$log_rt), 3),
    round(sd(rt_data_40_with$log_rt), 3),
    round(sd(rt_data_40_without$log_rt), 3)
  ),
  True_Structure = c("Random slopes + intercepts", "Fixed effect only",
                     "Random slopes + intercepts", "Fixed effect only")
)

knitr::kable(data_summaries,
             caption = "Four Test Datasets: 2×2 Design (Sample Size × Data Structure)",
             col.names = c("Scenario", "N", "Mean log-RT", "SD log-RT", "True Data Structure"),
             align = c("l", "r", "r", "r", "l"))
```

## Fit Models for All Four Scenarios

For each dataset, we'll fit two models:

1. **Simple model**: No random effects (just fixed effects) - `log_rt ~ condition`
2. **Complex model**: Random slopes for subjects - `(1 + condition | subject) + (1 | item)`

```{r fit-models}
#| cache: false

# Ensure brms is loaded
library(brms)

# Define priors
# Simple model: no random effects, just fixed effects + residual
rt_priors_simple <- c(
  prior(normal(6, 1.5), class = Intercept, lb = 4),
  prior(normal(0, 0.5), class = b),
  prior(exponential(1), class = sigma)
)

# Complex model: includes random effects
rt_priors_complex <- c(
  prior(normal(6, 1.5), class = Intercept, lb = 4),
  prior(normal(0, 0.5), class = b),
  prior(exponential(1), class = sigma),
  prior(exponential(1), class = sd),
  prior(lkj(2), class = cor)
)

# Create fits directory
dir.create("fits", showWarnings = FALSE, recursive = TRUE)

# Helper function to fit or load models
fit_or_load <- function(model_name, formula, data, priors) {
  filepath <- paste0("fits/", model_name, ".rds")
  if (file.exists(filepath)) {
    readRDS(filepath)
  } else {
    fit <- brm(
      formula = formula,
      data = data,
      family = gaussian(),
      prior = priors,
      chains = 4,
      iter = 2000,
      cores = 4,
      backend = "cmdstanr",
      seed = 1234,
      refresh = 0
    )
    saveRDS(fit, filepath)
    fit
  }
}

# Scenario 1: n=100, WITH random effects
fit_simple_100_with <- fit_or_load(
  "fit_simple_100_with",
  log_rt ~ condition,
  rt_data_100_with,
  rt_priors_simple
)

fit_complex_100_with <- fit_or_load(
  "fit_complex_100_with",
  log_rt ~ condition + (1 + condition | subject) + (1 | item),
  rt_data_100_with,
  rt_priors_complex
)

# Scenario 2: n=100, WITHOUT random effects
fit_simple_100_without <- fit_or_load(
  "fit_simple_100_without",
  log_rt ~ condition,
  rt_data_100_without,
  rt_priors_simple
)

fit_complex_100_without <- fit_or_load(
  "fit_complex_100_without",
  log_rt ~ condition + (1 + condition | subject) + (1 | item),
  rt_data_100_without,
  rt_priors_complex
)

# Scenario 3: n=40, WITH random effects
fit_simple_40_with <- fit_or_load(
  "fit_simple_40_with",
  log_rt ~ condition,
  rt_data_40_with,
  rt_priors_simple
)

fit_complex_40_with <- fit_or_load(
  "fit_complex_40_with",
  log_rt ~ condition + (1 + condition | subject) + (1 | item),
  rt_data_40_with,
  rt_priors_complex
)

# Scenario 4: n=40, WITHOUT random effects
fit_simple_40_without <- fit_or_load(
  "fit_simple_40_without",
  log_rt ~ condition,
  rt_data_40_without,
  rt_priors_simple
)

fit_complex_40_without <- fit_or_load(
  "fit_complex_40_without",
  log_rt ~ condition + (1 + condition | subject) + (1 | item),
  rt_data_40_without,
  rt_priors_complex
)
```



# Comparing Models with LOO Across Four Scenarios

## Add LOO Criterion to All Models

```{r add-loo}
# Add LOO criterion to all models
fit_simple_100_with <- add_criterion(fit_simple_100_with, "loo")
fit_complex_100_with <- add_criterion(fit_complex_100_with, "loo")

fit_simple_100_without <- add_criterion(fit_simple_100_without, "loo")
fit_complex_100_without <- add_criterion(fit_complex_100_without, "loo")

fit_simple_40_with <- add_criterion(fit_simple_40_with, "loo")
fit_complex_40_with <- add_criterion(fit_complex_40_with, "loo")

fit_simple_40_without <- add_criterion(fit_simple_40_without, "loo")
fit_complex_40_without <- add_criterion(fit_complex_40_without, "loo")
```

## Compare Models for Each Scenario

```{r compare-models}
# Compare models for each scenario
loo_comp_100_with <- loo_compare(fit_simple_100_with, fit_complex_100_with)
loo_comp_100_without <- loo_compare(fit_simple_100_without, fit_complex_100_without)
loo_comp_40_with <- loo_compare(fit_simple_40_with, fit_complex_40_with)
loo_comp_40_without <- loo_compare(fit_simple_40_without, fit_complex_40_without)

# Helper function to extract comparison info
extract_comparison <- function(loo_comp, scenario_name) {
  winner <- rownames(loo_comp)[1]
  winner_clean <- gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", winner)
  winner_clean <- tools::toTitleCase(winner_clean)
  
  elpd_diff <- loo_comp[2, "elpd_diff"]
  se_diff <- loo_comp[2, "se_diff"]
  ratio <- abs(elpd_diff) / se_diff
  
  if (ratio < 1) {
    interpretation <- "Essentially equivalent"
  } else if (ratio < 2) {
    interpretation <- "Weak evidence"
  } else if (ratio < 4) {
    interpretation <- "Moderate evidence"
  } else if (ratio < 10) {
    interpretation <- "Strong evidence"
  } else {
    interpretation <- "Very strong evidence"
  }
  
  data.frame(
    Scenario = scenario_name,
    Winner = winner_clean,
    ELPD_diff = round(elpd_diff, 2),
    SE_diff = round(se_diff, 2),
    Ratio = round(ratio, 2),
    Interpretation = interpretation
  )
}

# Create comprehensive comparison table
comparison_table <- rbind(
  extract_comparison(loo_comp_100_with, "n=100, WITH RE"),
  extract_comparison(loo_comp_100_without, "n=100, WITHOUT RE"),
  extract_comparison(loo_comp_40_with, "n=40, WITH RE"),
  extract_comparison(loo_comp_40_without, "n=40, WITHOUT RE")
)

knitr::kable(comparison_table,
             caption = "LOO Model Comparison Across Four Scenarios",
             col.names = c("Scenario", "Winner", "ELPD Δ", "SE", "Ratio", "Interpretation"),
             align = c("l", "l", "r", "r", "r", "l"),
             row.names = FALSE)
```

**Key patterns to observe:**

- **n=100, WITH RE**: Complex model wins decisively (ratio > 10)
- **n=100, WITHOUT RE**: Simple model wins weakly (ratio ≈ 1-2)
- **n=40, WITH RE**: Complex model wins strongly (ratio ≈ 6)
- **n=40, WITHOUT RE**: Models essentially equivalent (ratio < 1)

## Detailed Ratio Analysis

The ratio (|ELPD_diff| / SE) tells us how many standard errors separate the models. Here's a detailed breakdown:

```{r ratio-details}
# Extract detailed ratio information for both models in each scenario
ratio_details <- data.frame()

scenarios_list <- list(
  list(comp = loo_comp_100_with, name = "n=100, WITH RE"),
  list(comp = loo_comp_100_without, name = "n=100, WITHOUT RE"),
  list(comp = loo_comp_40_with, name = "n=40, WITH RE"),
  list(comp = loo_comp_40_without, name = "n=40, WITHOUT RE")
)

for (scenario in scenarios_list) {
  comp <- scenario$comp
  
  for (i in 1:nrow(comp)) {
    model_name <- rownames(comp)[i]
    model_clean <- tools::toTitleCase(gsub("fit_|_.*", "", model_name))
    
    elpd <- comp[i, "elpd_loo"]
    se_elpd <- comp[i, "se_elpd_loo"]
    elpd_diff <- comp[i, "elpd_diff"]
    se_diff <- comp[i, "se_diff"]
    
    if (i == 1) {
      # Best model
      ratio <- NA
      interpretation <- "Best model (reference)"
    } else {
      ratio <- abs(elpd_diff) / se_diff
      
      if (ratio < 1) {
        interpretation <- "Essentially equivalent"
      } else if (ratio < 2) {
        interpretation <- "Weak evidence"
      } else if (ratio < 4) {
        interpretation <- "Moderate evidence"
      } else if (ratio < 10) {
        interpretation <- "Strong evidence"
      } else {
        interpretation <- "Very strong evidence"
      }
    }
    
    ratio_details <- rbind(ratio_details, data.frame(
      Scenario = scenario$name,
      Model = model_clean,
      ELPD = round(elpd, 1),
      SE_ELPD = round(se_elpd, 1),
      ELPD_diff = ifelse(i == 1, "—", as.character(round(elpd_diff, 2))),
      SE_diff = ifelse(i == 1, "—", as.character(round(se_diff, 2))),
      Ratio = ifelse(i == 1, "—", as.character(round(ratio, 2))),
      Interpretation = interpretation
    ))
  }
}

knitr::kable(ratio_details,
             caption = "Detailed Model Comparison with Ratios (|ELPD_diff| / SE)",
             col.names = c("Scenario", "Model", "ELPD", "SE", "ELPD Δ", "SE Δ", "Ratio (SE)", "Interpretation"),
             align = c("l", "l", "r", "r", "r", "r", "r", "l"),
             row.names = FALSE)
```

**How to interpret the ratio:**

- **Ratio < 1**: Difference is smaller than the uncertainty → Models are essentially equivalent
- **Ratio 1-2**: Weak evidence → Difference is 1-2 standard errors, suggestive but not conclusive
- **Ratio 2-4**: Moderate evidence → Difference is 2-4 standard errors, reasonable confidence
- **Ratio 4-10**: Strong evidence → Difference is 4-10 standard errors, high confidence
- **Ratio > 10**: Very strong evidence → Difference is >10 standard errors, extremely confident

# Understanding ELPD

## What ELPD Actually Means

**ELPD** = "Expected Log Pointwise Predictive Density"

- **Expected**: We marginalize over all possible future data
- **Log**: Works on log scale for numerical stability
- **Pointwise**: Evaluated separately for each data point
- **Predictive Density**: How well the model predicts new data

**Key properties:**

- **Higher is better** (like R² in frequentist stats)
- **Difference matters**: Which model predicts new data better?
- **Not about fit to current data**: About generalization
- Takes into account the uncertainty of predictions

## Rule of Thumb for Model Comparison

**Interpreting `elpd_diff` (expected log pointwise predictive density difference):**

| elpd_diff | Ratio* | Interpretation | Action |
|-----------|--------|----------------|--------|
| \|diff\| < 4 | < 4 | Equivalent models | Pick simpler one |
| 4-10 | 4-10 | Moderate difference | Prefer larger elpd |
| > 10 | > 10 | Clear winner | Prefer larger elpd |

*Ratio = |elpd_diff| / se_diff (how many standard errors apart?)

**The ratio tells you:**

- < 1: Difference could be random noise
- 1-2: Weak evidence for difference
- 2-4: Moderate evidence
- > 4: Strong evidence

## Visualizations: Side-by-Side Comparisons

### Plot 1: ELPD Comparison (2×2 Grid)

```{r plot-loo-comparison-grid}
#| fig.width: 12
#| fig.height: 10

# Helper function to create ELPD plot for one scenario
plot_elpd_comparison <- function(loo_comp, scenario_name) {
  loo_df <- as.data.frame(loo_comp) %>%
    rownames_to_column("model") %>%
    mutate(
      model_clean = tools::toTitleCase(gsub("fit_|_.*", "", model)),
      is_winner = row_number() == 1
    )
  
  ggplot(loo_df, aes(x = elpd_loo, y = model_clean, color = is_winner)) +
    geom_pointrange(
      aes(xmin = elpd_loo - se_elpd_loo, 
          xmax = elpd_loo + se_elpd_loo),
      size = 1.2
    ) +
    scale_color_manual(values = c("FALSE" = "gray60", "TRUE" = "#E69F00")) +
    labs(
      title = scenario_name,
      x = "ELPD ± SE",
      y = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position = "none"
    )
}

# Create four plots
p1 <- plot_elpd_comparison(loo_comp_100_with, "n=100, WITH RE")
p2 <- plot_elpd_comparison(loo_comp_100_without, "n=100, WITHOUT RE")
p3 <- plot_elpd_comparison(loo_comp_40_with, "n=40, WITH RE")
p4 <- plot_elpd_comparison(loo_comp_40_without, "n=40, WITHOUT RE")

# Combine in 2×2 grid
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "ELPD Comparison Across Four Scenarios",
    subtitle = "Orange = Winner | Gray = Loser | Error bars show ±1 SE"
  )
```

### Plot 2: Pointwise ELPD Differences by RT (2×2 Grid)

```{r plot-elpd-differences-grid}
#| fig.width: 12
#| fig.height: 10

# Helper function to plot pointwise ELPD differences
plot_elpd_differences <- function(fit_simple, fit_complex, data, scenario_name) {
  diff_data <- data %>%
    mutate(
      diff_elpd = fit_complex$criteria$loo$pointwise[, "elpd_loo"] -
                  fit_simple$criteria$loo$pointwise[, "elpd_loo"],
      rt_category = case_when(
        rt < quantile(rt, 0.25) ~ "Fast",
        rt > quantile(rt, 0.75) ~ "Slow",
        TRUE ~ "Medium"
      ),
      rt_category = factor(rt_category, levels = c("Fast", "Medium", "Slow"))
    )
  
  ggplot(diff_data, aes(x = rt, y = diff_elpd, color = rt_category)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(alpha = 0.5, size = 1.5) +
    scale_color_manual(
      values = c("Fast" = "#009E73", "Medium" = "#56B4E9", "Slow" = "#E69F00"),
      name = "RT"
    ) +
    labs(
      title = scenario_name,
      x = "RT (ms)",
      y = "ELPD Diff\n(Complex - Simple)"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

# Create four plots
p1 <- plot_elpd_differences(fit_simple_100_with, fit_complex_100_with, 
                            rt_data_100_with, "n=100, WITH RE")
p2 <- plot_elpd_differences(fit_simple_100_without, fit_complex_100_without,
                            rt_data_100_without, "n=100, WITHOUT RE")
p3 <- plot_elpd_differences(fit_simple_40_with, fit_complex_40_with,
                            rt_data_40_with, "n=40, WITH RE")
p4 <- plot_elpd_differences(fit_simple_40_without, fit_complex_40_without,
                            rt_data_40_without, "n=40, WITHOUT RE")

# Combine in 2×2 grid
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Pointwise ELPD Differences: Complex vs Simple Model",
    subtitle = "Positive = Complex better | Negative = Simple better | Line at zero"
  )
```

**What to look for:**

- **WITH RE scenarios**: Positive values (complex better) when data truly has random slopes
- **WITHOUT RE scenarios**: Near zero or negative values (simple better or equivalent)
- **Sample size effect**: More scatter with n=40, clearer pattern with n=100

### Plot 3: Model Weights (2×2 Grid)

```{r plot-weights-grid}
#| fig.width: 12
#| fig.height: 10

# Helper function to plot model weights
plot_model_weights <- function(fit_simple, fit_complex, scenario_name) {
  weights <- model_weights(fit_simple, fit_complex, weights = "loo")
  weights_df <- data.frame(
    Model = c("Simple", "Complex"),
    Weight = as.numeric(weights)
  )
  
  ggplot(weights_df, aes(x = Model, y = Weight, fill = Model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = round(Weight, 3)), 
              vjust = -0.5, size = 3.5) +
    scale_fill_manual(values = c("Simple" = "#56B4E9", "Complex" = "#E69F00")) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(
      title = scenario_name,
      x = NULL,
      y = "Weight"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

# Create four plots
p1 <- plot_model_weights(fit_simple_100_with, fit_complex_100_with, "n=100, WITH RE")
p2 <- plot_model_weights(fit_simple_100_without, fit_complex_100_without, "n=100, WITHOUT RE")
p3 <- plot_model_weights(fit_simple_40_with, fit_complex_40_with, "n=40, WITH RE")
p4 <- plot_model_weights(fit_simple_40_without, fit_complex_40_without, "n=40, WITHOUT RE")

# Combine in 2×2 grid
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Model Weights: How Confident is LOO in Model Selection?",
    subtitle = "Weight ≈ 1.0 = Very confident | Weights ≈ 0.5 = Uncertain"
  )
```

### Plot 4: Pareto k Diagnostics (2×2 Grid)

```{r plot-pareto-k-grid}
#| fig.width: 12
#| fig.height: 10

# Helper function to plot Pareto k values
plot_pareto_k <- function(fit_complex, scenario_name) {
  k_values <- fit_complex$criteria$loo$diagnostics$pareto_k
  k_df <- data.frame(
    observation = 1:length(k_values),
    pareto_k = k_values,
    problematic = k_values > 0.7
  )
  
  ggplot(k_df, aes(x = observation, y = pareto_k, color = problematic)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange", alpha = 0.7) +
    geom_hline(yintercept = 0.7, linetype = "dashed", color = "red", alpha = 0.7) +
    scale_color_manual(values = c("FALSE" = "#56B4E9", "TRUE" = "#E69F00")) +
    labs(
      title = scenario_name,
      x = "Observation",
      y = "Pareto k"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

# Create four plots (use complex model for diagnostics)
p1 <- plot_pareto_k(fit_complex_100_with, "n=100, WITH RE")
p2 <- plot_pareto_k(fit_complex_100_without, "n=100, WITHOUT RE")
p3 <- plot_pareto_k(fit_complex_40_with, "n=40, WITH RE")
p4 <- plot_pareto_k(fit_complex_40_without, "n=40, WITHOUT RE")

# Combine in 2×2 grid
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Pareto k Diagnostics for Complex Model",
    subtitle = "k < 0.5 (good) | k < 0.7 (ok) | k > 0.7 (problematic)"
  )
```

# Cross-Validation Variants

Different CV strategies for different research questions:

**LOO (Leave-One-Out):**

- Default choice
- For general predictive performance
- Treats all observations as exchangeable

**K-fold CV:**

- Split data into K groups
- For multilevel models: sample from groups
- Useful for: predicting unseen data from existing subjects

**LOGO-CV (Leave-One-Group-Out):**

- Leave out entire groups (e.g., subjects)
- For predicting: completely new subjects
- More conservative than LOO

```{r cv-variants}
# Example: 10-fold cross-validation
# (computationally expensive - not run by default)
# fit_complex_100_with_kfold <- kfold(fit_complex_100_with, K = 10, folds = "stratified")

cv_methods <- data.frame(
  Method = c("loo()", "kfold()", "kfold() with group"),
  Description = c(
    "Leave-one-out (default)",
    "K-fold cross-validation",
    "Leave-one-group-out (e.g., subjects)"
  ),
  Use_Case = c(
    "General predictive performance",
    "Unseen data from existing subjects",
    "Completely new subjects"
  )
)

knitr::kable(cv_methods,
             caption = "Cross-Validation Variants in brms",
             col.names = c("Method", "Description", "Use Case"))
```

## Key Insights from Four Scenarios

```{r scenario-insights}
# Summarize key findings
insights_df <- data.frame(
  Scenario = c("n=100, WITH RE", "n=100, WITHOUT RE", "n=40, WITH RE", "n=40, WITHOUT RE"),
  Sample_Size = c("Large", "Large", "Small", "Small"),
  True_Structure = c("Complex", "Simple", "Complex", "Simple"),
  Expected_Winner = c("Complex", "Simple", "Complex", "Simple/Equiv"),
  Certainty = c("Very High", "Moderate", "High", "Low"),
  Key_Lesson = c(
    "LOO strongly identifies complexity",
    "LOO avoids overfitting (weak preference)",
    "Strong evidence even with less data",
    "Hard to distinguish with limited data"
  )
)

knitr::kable(insights_df,
             caption = "Summary of LOO Behavior Across Scenarios",
             col.names = c("Scenario", "Sample Size", "True Structure", "Expected Winner", "Certainty", "Key Lesson"),
             align = c("l", "l", "l", "l", "l", "l"))
```

**Main takeaways:**

1. **LOO works best with adequate data** (n ≥ 100): Clear winners, confident weights
2. **LOO respects true data structure**: Finds complexity when it exists, avoids it when it doesn't
3. **Small samples = high uncertainty**: Model weights closer to 0.5, wider error bars
4. **Pareto k generally good**: Few problematic observations across all scenarios

# Pareto k Diagnostics

## Identifying Influential Points

The LOO calculation uses Pareto Smoothed Importance Sampling (PSIS). The Pareto k diagnostic tells us if the approximation is reliable:

**Pareto k thresholds** (sample-size dependent):

- k < 0.5: Good (reliable estimate)
- 0.5 < k < 0.7: Okay (use with caution)
- k > 0.7: Bad (LOO estimate unreliable)

```{r check-pareto-k}
# Check Pareto k values for all scenarios
k_diagnostics <- data.frame()
k_threshold <- 0.7

scenarios <- list(
  list(fit = fit_complex_100_with, name = "n=100, WITH RE"),
  list(fit = fit_complex_100_without, name = "n=100, WITHOUT RE"),
  list(fit = fit_complex_40_with, name = "n=40, WITH RE"),
  list(fit = fit_complex_40_without, name = "n=40, WITHOUT RE")
)

for (scenario in scenarios) {
  k_values <- scenario$fit$criteria$loo$diagnostics$pareto_k
  n_bad <- sum(k_values > k_threshold)
  n_total <- length(k_values)
  
  status <- ifelse(n_bad > 0, 
                   "⚠ Consider reloo = TRUE",
                   "✓ All k values good")
  
  k_diagnostics <- rbind(k_diagnostics, data.frame(
    Scenario = scenario$name,
    Problematic = paste0(n_bad, " / ", n_total),
    Status = status
  ))
}

knitr::kable(k_diagnostics,
             caption = paste0("Pareto k Diagnostics Across Scenarios (threshold k > ", k_threshold, ")"),
             col.names = c("Scenario", "Observations with k > 0.7", "Status"),
             align = c("l", "r", "l"))
```

## Visualize Pareto k Values by Model and Scenario

```{r plot-pareto-k-by-model}
#| fig.width: 12
#| fig.height: 10

# Extract Pareto k values for all scenarios and both models
k_all_df <- bind_rows(
  # n=100, WITH RE
  data.frame(
    observation = 1:nrow(rt_data_100_with),
    k_simple = fit_simple_100_with$criteria$loo$diagnostics$pareto_k,
    k_complex = fit_complex_100_with$criteria$loo$diagnostics$pareto_k,
    scenario = "n=100, WITH RE"
  ),
  # n=100, WITHOUT RE
  data.frame(
    observation = 1:nrow(rt_data_100_without),
    k_simple = fit_simple_100_without$criteria$loo$diagnostics$pareto_k,
    k_complex = fit_complex_100_without$criteria$loo$diagnostics$pareto_k,
    scenario = "n=100, WITHOUT RE"
  ),
  # n=40, WITH RE
  data.frame(
    observation = 1:nrow(rt_data_40_with),
    k_simple = fit_simple_40_with$criteria$loo$diagnostics$pareto_k,
    k_complex = fit_complex_40_with$criteria$loo$diagnostics$pareto_k,
    scenario = "n=40, WITH RE"
  ),
  # n=40, WITHOUT RE
  data.frame(
    observation = 1:nrow(rt_data_40_without),
    k_simple = fit_simple_40_without$criteria$loo$diagnostics$pareto_k,
    k_complex = fit_complex_40_without$criteria$loo$diagnostics$pareto_k,
    scenario = "n=40, WITHOUT RE"
  )
) %>%
  pivot_longer(cols = starts_with("k_"), 
               names_to = "model", 
               values_to = "pareto_k") %>%
  mutate(
    model = factor(model, 
                   levels = c("k_simple", "k_complex"),
                   labels = c("Simple", "Complex")),
    scenario = factor(scenario, 
                      levels = c("n=100, WITH RE", "n=100, WITHOUT RE", 
                                "n=40, WITH RE", "n=40, WITHOUT RE"))
  )

# Create faceted plot
ggplot(k_all_df, aes(x = observation, y = pareto_k, color = model)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange", alpha = 0.7) +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "red", alpha = 0.7) +
  facet_grid(scenario ~ model, scales = "free_x") +
  scale_color_manual(values = c("Simple" = "#56B4E9", "Complex" = "#E69F00")) +
  labs(
    title = "Pareto k Diagnostics by Model and Scenario",
    subtitle = "Dashed lines: k = 0.5 (caution, orange) and k = 0.7 (problematic, red)",
    x = "Observation",
    y = "Pareto k",
    color = "Model"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
```

**What to look for:**

- Most points should be below 0.5 (good)
- Points between 0.5-0.7 (orange line) are okay but use with caution
- Points above 0.7 (red line) indicate unreliable LOO estimates
- Small sample scenarios (n=40) may show slightly higher k values due to limited data

## Handling Problematic Observations

If you have observations with k > 0.7:

```{r reloo-example}
# Example: Refit with exact LOO for problematic observations
# (only run if needed)

# Uncomment to run:
# loo_refit <- loo(fit_rt_complex, reloo = TRUE)

reloo_steps <- data.frame(
  Step = 1:4,
  Action = c(
    "Identify observations with k > 0.7",
    "Refit model leaving each one out exactly",
    "Use exact LOO for problematic observations",
    "Combine with PSIS-LOO for the rest"
  )
)

knitr::kable(reloo_steps,
             caption = "What reloo = TRUE Does",
             col.names = c("Step", "Action"),
             align = c("c", "l"))
```


## Influential Observations: Pareto k vs p_loo

The relationship between Pareto k and p_loo (effective number of parameters per observation) can reveal influential observations:

- **p_loo** measures how much each observation influences the model
- **High p_loo + high k**: Very influential observation that's hard to predict
- **Low p_loo + high k**: Outlier that doesn't strongly influence the model
- **High p_loo + low k**: Normal influential observation (e.g., high leverage point)

```{r plot-pareto-k-vs-p-loo}
#| fig.width: 12
#| fig.height: 10

# Helper function to plot pareto k vs p_loo
plot_k_vs_p_loo <- function(fit_complex, data, scenario_name) {
  loo_obj <- fit_complex$criteria$loo
  
  diag_df <- data.frame(
    observation = 1:nrow(data),
    pareto_k = loo_obj$diagnostics$pareto_k,
    p_loo = loo_obj$pointwise[, "p_loo"],
    problematic = loo_obj$diagnostics$pareto_k > 0.7 | loo_obj$pointwise[, "p_loo"] > 0.5
  )
  
  ggplot(diag_df, aes(x = pareto_k, y = p_loo, color = problematic)) +
    geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50", alpha = 0.7) +
    geom_vline(xintercept = 0.7, linetype = "dashed", color = "red", alpha = 0.5) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange", alpha = 0.5) +
    geom_point(aes(shape = problematic), alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("FALSE" = "#56B4E9", "TRUE" = "#E69F00")) +
    scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 19)) +
    labs(
      title = scenario_name,
      x = "Pareto k",
      y = "p_loo"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

# Create four plots
p1 <- plot_k_vs_p_loo(fit_complex_100_with, rt_data_100_with, "n=100, WITH RE")
p2 <- plot_k_vs_p_loo(fit_complex_100_without, rt_data_100_without, "n=100, WITHOUT RE")
p3 <- plot_k_vs_p_loo(fit_complex_40_with, rt_data_40_with, "n=40, WITH RE")
p4 <- plot_k_vs_p_loo(fit_complex_40_without, rt_data_40_without, "n=40, WITHOUT RE")

# Combine in 2×2 grid
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Pareto k vs p_loo Diagnostics (Complex Model)",
    subtitle = "Vertical lines: k = 0.5, 0.7 | Horizontal: p_loo = 0.5 | Orange = Problematic"
  )
```

**Interpretation:**

- Points in the **upper-right quadrant** (high k, high p_loo): Most concerning - influential outliers
- Points along the **right edge** (high k, low p_loo): Outliers with less model influence
- Points in the **upper-left** (low k, high p_loo): Normal high-leverage observations
- Most points should cluster in the **lower-left** (low k, low p_loo): Well-behaved observations

# Comparing WAIC and LOO

## Understanding the Differences

Both WAIC and LOO estimate out-of-sample predictive accuracy, but they use different approaches:

**WAIC (Watanabe-Akaike Information Criterion):**

- **Method**: Uses the entire dataset at once
- **Approximation**: Based on asymptotic theory (assumes large sample sizes)
- **p_waic**: Estimates effective number of parameters from posterior variance
- **Pros**: Fast to compute, simple formula
- **Cons**: Can be unstable with small samples or influential observations, no diagnostics

**LOO-PSIS (Leave-One-Out with Pareto Smoothed Importance Sampling):**

- **Method**: Simulates leaving each observation out one at a time
- **Approximation**: Uses importance sampling (no asymptotic assumptions needed)
- **p_loo**: Estimates effective parameters from LOO differences
- **Pros**: More stable, includes diagnostics (Pareto k), works better with small samples
- **Cons**: Slightly slower (but still fast with PSIS)

**Key technical differences:**

| Aspect | WAIC | LOO |
|--------|------|-----|
| Estimation | Posterior variance | Importance sampling |
| Diagnostics | None | Pareto k values |
| Small samples | Can be unstable | More robust |
| Influential obs | No warning | Flags with high k |
| Computation | Slightly faster | Fast enough |

**When they disagree:**

- Different rankings suggest **influential observations** or **model instability**
- Check Pareto k diagnostics - high k values indicate LOO is more reliable
- WAIC may overestimate predictive accuracy when observations are very influential

**Recommendation**: Use LOO by default. The Pareto k diagnostics are invaluable for catching problems.

## Computing Both Criteria

```{r compare-waic-loo}
# Add WAIC criterion to all models
fit_simple_100_with <- add_criterion(fit_simple_100_with, "waic")
fit_complex_100_with <- add_criterion(fit_complex_100_with, "waic")

fit_simple_100_without <- add_criterion(fit_simple_100_without, "waic")
fit_complex_100_without <- add_criterion(fit_complex_100_without, "waic")

fit_simple_40_with <- add_criterion(fit_simple_40_with, "waic")
fit_complex_40_with <- add_criterion(fit_complex_40_with, "waic")

fit_simple_40_without <- add_criterion(fit_simple_40_without, "waic")
fit_complex_40_without <- add_criterion(fit_complex_40_without, "waic")

# Compare with WAIC for each scenario
waic_comp_100_with <- loo_compare(fit_simple_100_with, fit_complex_100_with, criterion = "waic")
waic_comp_100_without <- loo_compare(fit_simple_100_without, fit_complex_100_without, criterion = "waic")
waic_comp_40_with <- loo_compare(fit_simple_40_with, fit_complex_40_with, criterion = "waic")
waic_comp_40_without <- loo_compare(fit_simple_40_without, fit_complex_40_without, criterion = "waic")

# Create comparison table
ranking_comparison <- data.frame(
  Scenario = c("n=100, WITH RE", "n=100, WITHOUT RE", "n=40, WITH RE", "n=40, WITHOUT RE"),
  WAIC_Winner = c(
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(waic_comp_100_with)[1]),
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(waic_comp_100_without)[1]),
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(waic_comp_40_with)[1]),
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(waic_comp_40_without)[1])
  ),
  LOO_Winner = c(
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(loo_comp_100_with)[1]),
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(loo_comp_100_without)[1]),
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(loo_comp_40_with)[1]),
    gsub("fit_|_100_with|_100_without|_40_with|_40_without", "", rownames(loo_comp_40_without)[1])
  )
) %>%
  mutate(
    WAIC_Winner = tools::toTitleCase(WAIC_Winner),
    LOO_Winner = tools::toTitleCase(LOO_Winner),
    Agreement = ifelse(WAIC_Winner == LOO_Winner, "✓ Agree", "✗ Differ")
  )

knitr::kable(ranking_comparison,
             caption = "Model Rankings: WAIC vs LOO Across Scenarios",
             col.names = c("Scenario", "WAIC Winner", "LOO Winner", "Agreement"),
             align = c("l", "l", "l", "c"))
```

**Interpreting agreement/disagreement:**

- **Rankings identical**: Both methods agree - conclusions are robust
- **Small differences in values**: Normal - both methods have uncertainty
- **Rankings differ**: Investigate! Check Pareto k diagnostics and look for influential observations

# Summary and Best Practices

## When to Use LOO

**✅ Use LOO for:**

- Comparing model structures (e.g., with/without random slopes)
- Feature selection (which predictors to include?)
- Comparing different likelihoods (Gaussian vs. Student-t)
- Choosing between regularizing vs. non-regularizing priors

**❌ Don't use LOO for:**

- Testing specific scientific hypotheses (use posterior distributions)
- Comparing very similar models (differences may not be meaningful)
- With very small samples (k < 20 per group)

## Workflow Recommendations

**Step 1: Prior sensitivity analysis**

- Fit same model with different reasonable priors
- Check if conclusions are robust
- See `04_comparing_priors_rt.qmd`

**Step 2: Model comparison with LOO**

- Compare different model structures
- Choose best model using ELPD difference
- Check Pareto k diagnostics

**Step 3: Final model validation**

- Posterior predictive checks
- Check convergence diagnostics
- Report model comparison results

## Reporting LOO Results

**Minimal reporting:**

```
We compared three models using LOO-CV: simple (random intercepts only),
complex (random slopes for subjects), and full (random slopes for both 
subjects and items). The complex model showed the best predictive 
performance (ELPD = 105.2), outperforming the simple model by 5.2 
(SE = 2.1, ratio = 2.5). All Pareto k values were < 0.5, indicating 
reliable estimates.
```

**Complete reporting:**

- ELPD differences and standard errors
- Model weights
- Pareto k diagnostics
- Interpretation of ratios
- Reasoning for final model choice

## Session Info

```{r session-info}
sessionInfo()
```
