---
title: "6: Sequential Testing: ROPE vs Bayes Factor vs LOO (Continuous)"
subtitle: "Understanding Divergences in Bayesian Decision Making"
date: today
author: "Job Schepens"
format: 
  html:
    toc: true
    toc-depth: 5
    toc-expand: 5
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    keep-md: true
    embed-resources: false
    link-external-newwindow: true
    link-external-icon: true
    theme:
      light: flatly
      dark: darkly
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    documentclass: article
    fontsize: 11pt
    geometry: "top=1in,bottom=1in,left=0.75in,right=0.75in"
    echo: false
    keep-md: false
execute:
  freeze: auto
  cache: true
  warning: false
  message: false
---


```{r}
#| label: setup
#| message: false
#| cache: false

library(brms)
library(tidyverse)
library(bayesplot)
library(posterior)
library(bayestestR)
library(emmeans)
library(marginaleffects)
library(patchwork)
library(tidybayes)

theme_set(theme_minimal(base_size = 14))
```

# Part 2: Continuous Predictor Simulation

In this second part, we repeat the sequential testing process for a **continuous predictor** (e.g., word frequency or trial number) to see if decision metrics behave similarly.

## Data Generation (Continuous)

We simulate a log-RT variable with a continuous predictor $X$ (centered, range approx -2 to 2).
True Effect of X: 0.05 (approx 5% change per unit of X).

```{r}
#| label: data-generation-cont

set.seed(999)

generate_data_cont <- function(n_subj) {
  n_item <- 10
  
  expand_grid(
    subject = factor(1:n_subj),
    item = factor(1:n_item),
    trial_x = seq(-1, 1, length.out = 10) # Continuous predictor per item/trial
  ) %>%
    group_by(subject) %>%
    mutate(
      subj_intercept = rnorm(1, 0, 0.15),
      subj_slope = rnorm(1, 0, 0.05)
    ) %>%
    ungroup() %>%
    mutate(
      # True effect: 0.08 per unit of X
      log_rt = 6.0 + 
               subj_intercept + 
               0.08 * trial_x + 
               trial_x * subj_slope + 
               rnorm(n(), 0, 0.2),
      y = log_rt
    ) %>%
    select(subject, item, trial_x, y)
}
```

## Simulation Loop (Continuous)

We fit three models again:
1. **Null**: `y ~ 1 + (1|subject) + (1|item)`
2. **Wide**: `y ~ trial_x`, prior `normal(0, 5)`
3. **Narrow**: `y ~ trial_x`, prior `normal(0, 0.2)`

```{r}
#| label: simulation-loop-cont
#| output: false
#| cache: true

# Same extended sample sizes
sample_sizes_cont <- c(1:10, seq(15, 30, 5), 40, 50, 100)
results_list_cont <- list()

# Generate full dataset
max_n_cont <- max(sample_sizes_cont)
data_full_cont <- generate_data_cont(max_n_cont)

for (n in sample_sizes_cont) {
  
  # Subset Data
  data_n <- data_full_cont %>% 
    filter(as.integer(subject) <= n) %>% 
    droplevels()
    
  # Fit Models
  # H0: Null (Random slopes for X could be included, but keeping simple for null)
  fit_null <- brm(
    y ~ 1 + (1 + trial_x | subject) + (1 | item),
    data = data_n,
    prior = c(
      prior(normal(6, 0.5), class = Intercept),
      prior(exponential(2), class = sd),
      prior(exponential(2), class = sigma)
    ),
    file = paste0("models/fit_cont_null_N", n),
    silent = 2, refresh = 0, seed = 123
  )
  
  # H1 Wide
  fit_wide <- brm(
    y ~ trial_x + (1 + trial_x | subject) + (1 | item),
    data = data_n,
    prior = c(
      prior(normal(6, 0.5), class = Intercept),
      prior(normal(0, 1.0), class = b), # Wide for slope
      prior(exponential(2), class = sd),
      prior(exponential(2), class = sigma)
    ),
    sample_prior = "yes",
    file = paste0("models/fit_cont_wide_N", n),
    silent = 2, refresh = 0, seed = 123
  )
  
  # H1 Narrow
  fit_narrow <- brm(
    y ~ trial_x + (1 + trial_x | subject) + (1 | item),
    data = data_n,
    prior = c(
      prior(normal(6, 0.5), class = Intercept),
      prior(normal(0, 0.1), class = b), # Narrow expectation for slope
      prior(exponential(2), class = sd),
      prior(exponential(2), class = sigma)
    ),
    sample_prior = "yes",
    file = paste0("models/fit_cont_narrow_N", n),
    silent = 2, refresh = 0, seed = 123
  )
  
  # Metrics
  
  # ROPE [-0.05, 0.05] on slope
  rope_wide_df <- rope(fit_wide, range = c(-0.05, 0.05), ci = 0.95)
  rope_narrow_df <- rope(fit_narrow, range = c(-0.05, 0.05), ci = 0.95)
  
  # Select duplicate rows (Intercept + Trial_x) -> Take only trial_x (usually 2nd param)
  # Safer: filter by Parameter name if possible, or take index 2
  rope_wide <- rope_wide_df$ROPE_Percentage[2]
  rope_narrow <- rope_narrow_df$ROPE_Percentage[2]
  
  # Bayes Factor (Slope = 0)
  bf_wide <- 1 / hypothesis(fit_wide, "trial_x = 0")$hypothesis$Evid.Ratio
  bf_narrow <- 1 / hypothesis(fit_narrow, "trial_x = 0")$hypothesis$Evid.Ratio
  
  # LOO
  loo_null <- loo(fit_null)
  loo_wide <- loo(fit_wide)
  loo_narrow <- loo(fit_narrow)
  
  # Calculate LOO gain and SE of the difference
  diff_wide <- loo_wide$pointwise[,"elpd_loo"] - loo_null$pointwise[,"elpd_loo"]
  gain_wide <- sum(diff_wide)
  se_wide <- sqrt(length(diff_wide) * var(diff_wide))
  
  diff_narrow <- loo_narrow$pointwise[,"elpd_loo"] - loo_null$pointwise[,"elpd_loo"]
  gain_narrow <- sum(diff_narrow)
  se_narrow <- sqrt(length(diff_narrow) * var(diff_narrow))
  
  # Estimates
  est_narrow <- fixef(fit_narrow, probs = c(0.025, 0.975))["trial_x", ]
  est_wide <- fixef(fit_wide, probs = c(0.025, 0.975))["trial_x", ]
  
  results_list_cont[[paste0("N", n)]] <- tibble(
    N = n,
    # Est
    Est_Narrow = as.numeric(est_narrow["Estimate"]), 
    Low_Narrow = as.numeric(est_narrow["Q2.5"]), 
    High_Narrow = as.numeric(est_narrow["Q97.5"]),
    Est_Wide = as.numeric(est_wide["Estimate"]), 
    Low_Wide = as.numeric(est_wide["Q2.5"]), 
    High_Wide = as.numeric(est_wide["Q97.5"]),
    # Metrics
    ROPE_in_prob_Narrow = as.numeric(rope_narrow), 
    ROPE_in_prob_Wide = as.numeric(rope_wide),
    BF10_Narrow = as.numeric(bf_narrow), 
    BF10_Wide = as.numeric(bf_wide),
    LOO_gain_Narrow = as.numeric(gain_narrow), 
    LOO_se_Narrow = as.numeric(se_narrow),
    LOO_gain_Wide = as.numeric(gain_wide),
    LOO_se_Wide = as.numeric(se_wide)
  )
}

results_df_cont <- bind_rows(results_list_cont)
```

## Continuous Results Table

```{r}
#| label: results-table-cont
#| cache: true
results_df_cont %>%
  pivot_longer(
    cols = -N,
    names_to = c("Metric", "Prior"),
    names_pattern = "(.*)_(Narrow|Wide)"
  ) %>%
  mutate(value = map_dbl(value, ~ if(is.list(.x)) as.numeric(.x[[1]]) else as.numeric(.x))) %>%
  pivot_wider(names_from = Metric, values_from = value) %>%
  unnest(cols = everything()) %>%
  mutate(Estimate_CI = sprintf("%.2f [%.2f, %.2f]", Est, Low, High)) %>%
  select(N, Prior, Estimate_CI, BF10, ROPE_Prob = ROPE_in_prob, LOO_Gain = LOO_gain) %>%
  arrange(N, desc(Prior)) %>%
  knitr::kable(digits = 3, caption = "Continuous Predictor: Decision Metrics")
```

## Visualization (Continuous)

```{r}
#| label: wiz-cont
#| fig.width: 10
#| fig.height: 8

# First, ensure all columns are properly numeric by unnesting and converting
results_clean <- results_df_cont %>%
  mutate(across(everything(), ~ {
    if(is.list(.x)) {
      map_dbl(.x, ~ if(length(.x) > 0) as.numeric(.x[[1]]) else NA_real_)
    } else {
      as.numeric(.x)
    }
  })) %>%
  # Double-check: convert to data frame to remove any remaining list structure
  as.data.frame() %>%
  as_tibble()

# 1. Estimates
p_est_c <- results_clean %>%
  select(N, starts_with("Est"), starts_with("Low"), starts_with("High")) %>%
  pivot_longer(cols = -N, names_to = c("Type", "Prior"), names_sep = "_") %>%
  pivot_wider(names_from = Type, values_from = value) %>%
  unnest(cols = c(Est, Low, High)) %>%
  ggplot(aes(x = factor(N), y = Est, color = Prior, group = Prior)) +
  geom_pointrange(aes(ymin = Low, ymax = High), position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = 0.08, linetype = "dashed", color = "gray50") +
  labs(title = "Estimates (Slope)", y = "Beta") +
  theme(legend.position = "bottom")

# 2. Metrics
plot_met_c <- results_clean %>%
  select(N, contains("BF10"), contains("ROPE"), contains("LOO")) %>%
  pivot_longer(cols = -N, names_to = c("MetricType", "Prior"), names_pattern = "(.*)_(Wide|Narrow)", values_to = "Value") %>%
  mutate(Value = as.numeric(Value))

p_bf_c <- plot_met_c %>% filter(MetricType == "BF10") %>%
  ggplot(aes(x = factor(N), y = Value, color = Prior, group = Prior)) +
  geom_line() + geom_point() + scale_y_log10() +
  geom_hline(yintercept = c(1, 3), linetype = c("dashed", "dotted")) +
  labs(title = "Bayes Factor (Slope != 0)")

p_rope_c <- plot_met_c %>% filter(MetricType == "ROPE_in_prob") %>%
  ggplot(aes(x = factor(N), y = Value, color = Prior, group = Prior)) +
  geom_line() + geom_point() +
  labs(title = "ROPE Prob (Slope ~ 0)")

p_loo_c <- plot_met_c %>% filter(MetricType == "LOO_gain") %>%
  ggplot(aes(x = factor(N), y = Value, color = Prior, group = Prior)) +
  geom_line() + geom_point() + geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "LOO Gain")

(p_est_c + p_bf_c + p_rope_c + p_loo_c) + 
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

```

## Visualization (Zoomed: N >= 10)

Checking convergence by ignoring small N chaos.

```{r}
#| label: wiz-cont-zoomed
#| fig.width: 10
#| fig.height: 8

results_clean_zoom <- results_clean %>% filter(N >= 10)

# 1. Estimates
p_est_cz <- results_clean_zoom %>%
  select(N, starts_with("Est"), starts_with("Low"), starts_with("High")) %>%
  pivot_longer(cols = -N, names_to = c("Type", "Prior"), names_sep = "_") %>%
  pivot_wider(names_from = Type, values_from = value) %>%
  unnest(cols = c(Est, Low, High)) %>%
  ggplot(aes(x = factor(N), y = Est, color = Prior, group = Prior)) +
  geom_pointrange(aes(ymin = Low, ymax = High), position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = 0.08, linetype = "dashed", color = "gray50") +
  labs(title = "Estimates (Slope)", y = "Beta", subtitle="N >= 10") +
  theme(legend.position = "bottom")

# 2. Metrics
plot_met_cz <- results_clean_zoom %>%
  select(N, contains("BF10"), contains("ROPE"), contains("LOO_gain")) %>% # Exclude SE for this long format
  pivot_longer(cols = -N, names_to = c("MetricType", "Prior"), names_pattern = "(.*)_(Wide|Narrow)", values_to = "Value") %>%
  mutate(Value = as.numeric(Value))

p_bf_cz <- plot_met_cz %>% filter(MetricType == "BF10") %>%
  ggplot(aes(x = factor(N), y = Value, color = Prior, group = Prior)) +
  geom_line() + geom_point() + scale_y_log10() +
  geom_hline(yintercept = c(1, 3), linetype = c("dashed", "dotted")) +
  labs(title = "Bayes Factor (Slope != 0)")

p_rope_cz <- plot_met_cz %>% filter(MetricType == "ROPE_in_prob") %>%
  ggplot(aes(x = factor(N), y = Value, color = Prior, group = Prior)) +
  geom_line() + geom_point() +
  labs(title = "ROPE Prob (Slope ~ 0)")

p_loo_cz <- plot_met_cz %>% filter(MetricType == "LOO_gain") %>%
  ggplot(aes(x = factor(N), y = Value, color = Prior, group = Prior)) +
  geom_line() + geom_point() + geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "LOO Gain")

(p_est_cz + p_bf_cz + p_rope_cz + p_loo_cz) + 
  plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Visualization (Uncertainty: N >= 10)

Including LOO uncertainty ($SE_{diff}$).

```{r}
#| label: wiz-cont-uncertainty
#| fig.width: 10
#| fig.height: 8

# Create specific LOO data frame with SE
plot_loo_unc_c <- results_clean_zoom %>%
  select(N, LOO_gain_Wide, LOO_se_Wide, LOO_gain_Narrow, LOO_se_Narrow) %>%
  pivot_longer(
    cols = -N,
    names_to = c(".value", "Prior"),
    names_pattern = "LOO_(.*)_(Wide|Narrow)"
  ) %>% # Creates N, Prior, gain, se
  mutate(
    # Unnest just in case, though results_clean should be flat
    gain = as.numeric(gain),
    se = as.numeric(se),
    Low = gain - 2*se,
    High = gain + 2*se
  )

p_loo_unc_c <- ggplot(plot_loo_unc_c, aes(x = factor(N), y = gain, color = Prior, group = Prior)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_ribbon(aes(ymin = Low, ymax = High, fill = Prior), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) + 
  geom_point(size = 3) +
  labs(title = "Predictive Gain (LOO)", subtitle = "Mean +/- 2 SE", y = "Delta elpd")

(p_est_cz + p_bf_cz + p_rope_cz + p_loo_unc_c) + 
  plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r}
#| label: lmm-cont-sim
#| cache: true
#| message: false
#| warning: false

# Ensure cache directory exists
library(lme4)
library(brms)
dir.create("cache", showWarnings = FALSE)
results_cont_path <- "cache/sim_cont_results.rds"
dir.create("models/cache_fits", recursive = TRUE, showWarnings = FALSE)

# Pre-compile a template model (Standard/Wide Prior) once to save C++ time
# We use sample_prior="yes" for Savage-Dickey BF
if (!file.exists("models/brm_template_cont.rds")) {
  # Minimal data for template
  tmp_dat <- tibble(subject=factor(1:2), item=factor(1:2), trial_x=c(-1,1), y=c(0,0))
  prior_h1 <- c(
    prior(normal(6, 0.5), class = Intercept),
    prior(normal(0, 1.0), class = b, coef = "trial_x"), # Wide H1
    prior(exponential(2), class = sd),
    prior(exponential(2), class = sigma)
  )
  brm_template <- brm(
    y ~ trial_x + (1|subject) + (1|item),
    data = tmp_dat,
    prior = prior_h1,
    sample_prior = "yes",
    chains = 2, iter = 2000,
    file = "models/brm_template_cont",
    silent = 2, refresh = 0, seed = 123
  )
} else {
  brm_template <- readRDS("models/brm_template_cont.rds")
}

run_cont_simulation <- function(n_sim = 20, true_effect = 0, noise_sd = 0.2, cache_prefix = "sim_cont", brm_model = NULL) {
  
  # Ensure cache dir exists
  dir.create("results_cont", showWarnings = FALSE)
  
  results <- tibble()
  
  for (i in 1:n_sim) {
    # File-based cache for THIS simulation iteration
    cache_file <- paste0("results_cont/", cache_prefix, "_iter", i, ".rds")
    
    # 1. Load partial progress if exists
    iter_results <- tibble()
    if (file.exists(cache_file)) {
      iter_results <- readRDS(cache_file)
    }
    
    # 2. Determine pending checks
    # If file exists but is incomplete, we resume.
    done_n <- numeric(0)
    if (nrow(iter_results) > 0) {
      done_n <- unique(iter_results$check_n)
    }
    
    check_points <- seq(15, 60, by = 5)
    points_to_run <- setdiff(check_points, done_n)
    
    if (length(points_to_run) == 0) {
      # Already complete for this iteration
      results <- bind_rows(results, iter_results)
      next
    }
    
    # 3. Generate Data (Fixed Seed/State for Reproducibility across resumes?)
    # Generating full dataset 100% freshly might drift if set.seed isn't managed per iter.
    # ideally we save the dataset too? Or imply robust reproducibility via seed?
    # For now, we regenerate. Note: if reusing partial results, we MUST ensure the dataset 
    # for the NEW steps is identical to the OLD steps.
    # Quick Fix: Set seed based on iteration ID to guarantee consistency.
    set.seed(999 + i) 
    
    n_param <- 100
    dat_full <- expand_grid(subject=factor(1:n_param), item=factor(1:10), trial_x=seq(-1,1,length.out=10)) %>%
      group_by(subject) %>% mutate(subj_int = rnorm(1,0,0.15)) %>% ungroup() %>%
      mutate(y = 6 + subj_int + true_effect*trial_x + rnorm(n(), 0, noise_sd))
    
    # Initialize stopping flags based on PREVIOUS partial results if strictly sequential
    # If we stopped at N=30, we must carry that state.
    stopped_p <- FALSE
    stopped_bf10 <- FALSE
    stopped_bf10_brms <- FALSE
    
    if(nrow(iter_results) > 0) {
        stopped_p <- any(iter_results$decision_p)
        stopped_bf10 <- any(iter_results$decision_bf10)
        if("decision_bf10_brms" %in% names(iter_results)) {
          stopped_bf10_brms <- any(iter_results$decision_bf10_brms, na.rm=TRUE)
        }
    }
    
    points_to_run <- sort(points_to_run)
    
    for (n in points_to_run) {
      curr_dat <- dat_full %>% filter(as.numeric(subject) <= n)
      
      # A. Fast Frequentist + BIC Approx
      ctrl <- lmerControl(calc.derivs = FALSE)
      
      m1 <- suppressMessages(lmer(y ~ trial_x + (1 + trial_x|subject) + (1|item), 
                                data = curr_dat, REML=FALSE, control = ctrl))
      m0 <- suppressMessages(lmer(y ~ 1 + (1 + trial_x|subject) + (1|item), 
                                data = curr_dat, REML=FALSE, control = ctrl))
      
      # Stats from LogLik
      ll1 <- as.numeric(logLik(m1))
      ll0 <- as.numeric(logLik(m0))
      df1 <- attr(logLik(m1), "df")
      df0 <- attr(logLik(m0), "df")
      
      chisq_val <- 2 * (ll1 - ll0)
      p_val <- pchisq(chisq_val, df = df1 - df0, lower.tail = FALSE)
      if(is.na(p_val)) p_val <- 1
      
      n_obs <- nrow(curr_dat)
      bic1 <- df1 * log(n_obs) - 2 * ll1
      bic0 <- df0 * log(n_obs) - 2 * ll0
      bf10 <- exp((bic0 - bic1) / 2)
      
      # B. brms (Full Bayesian)
      bf10_brms <- NA # Default if fails
      if (!is.null(brm_model)) {
        
        # Optimize: Store brms models to disk?
        # User requested saving models. We can do it!
        model_file <- paste0("models/cache_fits/", cache_prefix, "_iter", i, "_N", n)
        
        # Check if model exists? brms has 'file' arg, but update() logic is manual.
        # We'll use update() with 'file' argument if possible, or save manually.
        # update() with file= writes the updated model to file.
        
        # NOTE: update() with 'file' might assume the file contains the model to update FROM, 
        # or the location to save TO? brms::update usually just takes the object. 
        # Actually brms::update returns a NEW object. If we want caching, we use 'file'.
        
        m_brms <- update(brm_model, newdata = curr_dat, 
                         chains = 2, iter = 2000, 
                         silent = 2, refresh = 0,
                         file = model_file) # Persist model!
        
        # Test trial_x = 0
        h_res <- try(hypothesis(m_brms, "trial_x = 0"), silent = TRUE)
        if (!inherits(h_res, "try-error")) {
             bf10_brms <- 1 / h_res$hypothesis$Evid.Ratio
        }
      }
      
      # Decisions
      if (!stopped_p && p_val < 0.05) stopped_p <- TRUE
      if (!stopped_bf10 && bf10 > 10) stopped_bf10 <- TRUE
      if (!stopped_bf10_brms && !is.na(bf10_brms) && bf10_brms > 10) stopped_bf10_brms <- TRUE
      
      new_row <- tibble(
        sim_id = i,
        check_n = n,
        decision_p = stopped_p,
        decision_bf10 = stopped_bf10,
        decision_bf10_brms = stopped_bf10_brms
      )
      
      iter_results <- bind_rows(iter_results, new_row)
      
      # Save IMMEDIATELY after each step
      saveRDS(iter_results, cache_file)
    }
    
    results <- bind_rows(results, iter_results)
  }
  return(results)
}

if (file.exists(results_cont_path)) {
  cont_summary <- readRDS(results_cont_path)
} else {
  # Run 4 Simulation Conditions with brms template
  # We reuse the same template for all since the model FORMULA is the same (H1 model)
  # and we just update data.
  sim_cont_low_h0 <- run_cont_simulation(n_sim = 20, true_effect = 0, noise_sd = 0.2, 
                                         cache_prefix = "low_h0_brms", brm_model = brm_template) %>%
    mutate(Noise = "Low Noise (SD=0.2)", Scenario = "H0 (No Effect)")
  
  sim_cont_low_h1 <- run_cont_simulation(n_sim = 20, true_effect = 0.05, noise_sd = 0.2, 
                                         cache_prefix = "low_h1_brms", brm_model = brm_template) %>%
    mutate(Noise = "Low Noise (SD=0.2)", Scenario = "H1 (Effect=0.05)")
  
  sim_cont_high_h0 <- run_cont_simulation(n_sim = 20, true_effect = 0, noise_sd = 0.8, 
                                          cache_prefix = "high_h0_brms", brm_model = brm_template) %>%
    mutate(Noise = "High Noise (SD=0.8)", Scenario = "H0 (No Effect)")
  
  sim_cont_high_h1 <- run_cont_simulation(n_sim = 20, true_effect = 0.05, noise_sd = 0.8, 
                                          cache_prefix = "high_h1_brms", brm_model = brm_template) %>%
    mutate(Noise = "High Noise (SD=0.8)", Scenario = "H1 (Effect=0.05)")
  
  # Combine
  cont_data_all <- bind_rows(sim_cont_low_h0, sim_cont_low_h1, sim_cont_high_h0, sim_cont_high_h1)
  
  cont_summary <- cont_data_all %>%
    group_by(Noise, Scenario, check_n) %>%
    summarise(
      prop_p = mean(decision_p),
      prop_bf10 = mean(decision_bf10),
      prop_bf10_brms = mean(decision_bf10_brms),
      .groups = "drop"
    ) %>%
    pivot_longer(cols = starts_with("prop"), names_to = "Method", values_to = "Rate")
  
  saveRDS(cont_summary, results_cont_path)
}
```

## Simulation Results Visualization

```{r}
#| label: lmm-cont-plot
#| fig.width: 12
#| fig.height: 10

if (!exists("cont_summary")) {
  if (file.exists("cache/sim_cont_results.rds")) {
    cont_summary <- readRDS("cache/sim_cont_results.rds")
  } else {
    stop("Continuous simulation results not found. Run previous chunk.")
  }
}

# 4-Panel Plot Helper
plot_panel <- function(data, title) {
  ggplot(data, aes(x = check_n, y = Rate, color = Method, group = Method)) + 
    # Use position dodge to avoid overlap
    geom_line(size = 1.2, position = position_dodge(width = 1.5), alpha = 0.8) + 
    geom_point(size = 3, position = position_dodge(width = 1.5), alpha = 0.8) +
    ylim(0, 1) + 
    labs(title = title, x = NULL) 
}

p1 <- cont_summary %>% filter(Noise == "Low Noise (SD=0.2)", Scenario == "H0 (No Effect)") %>%
  plot_panel("False Positives (Low Noise)")

p2 <- cont_summary %>% filter(Noise == "Low Noise (SD=0.2)", Scenario == "H1 (Effect=0.05)") %>%
  plot_panel("Power (Low Noise)")

p3 <- cont_summary %>% filter(Noise == "High Noise (SD=0.8)", Scenario == "H0 (No Effect)") %>%
  plot_panel("False Positives (High Noise)") + labs(x="Sample Size (N)")

p4 <- cont_summary %>% filter(Noise == "High Noise (SD=0.8)", Scenario == "H1 (Effect=0.05)") %>%
  plot_panel("Power (High Noise)") + labs(x="Sample Size (N)")

(p1 + p2) / (p3 + p4) + 
  plot_layout(guides = "collect") & 
  scale_color_manual(values = c("prop_p"="#E74C3C", "prop_bf10"="#2ECC71", "prop_bf10_brms"="#9B59B6"),
                     labels = c("prop_p"="P-Value < .05", "prop_bf10"="BIC BF > 10", "prop_bf10_brms"="brms BF > 10")) &
  theme(legend.position = "bottom")
```

## Simulation Results Table

```{r}
#| label: results-table-cont-sim
#| echo: false

cont_summary %>%
  pivot_wider(names_from = Method, values_from = Rate) %>%
  mutate(across(starts_with("prop"), ~ sprintf("%.1f%%", .x * 100))) %>%
  rename(
    `P-Value (<.05)` = prop_p,
    `Bayes Factor (BIC) > 10` = prop_bf10,
    `Bayes Factor (brms) > 10` = prop_bf10_brms
  ) %>%
  knitr::kable(caption = "Sequential Testing Results: Continuous Predictor Simulation")
```
