---
title: "BRMS Workshop: Basics of Bayesian Regression"
author: "BRMS Workshop"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)

# Suppress Stan messages for cleaner output
options(mc.cores = parallel::detectCores())

# Set BRMS backend (cmdstanr should be pre-compiled in container)
# If not available, will fall back to rstan
if (requireNamespace("cmdstanr", quietly = TRUE)) {
  options(brms.backend = "cmdstanr")
  cat("✓ Using CmdStan backend (pre-compiled)\n")
} else {
  cat("! CmdStan backend not available, using RStan\n")
}
```

# Introduction to BRMS

Welcome to the Bayesian Regression Models using Stan (BRMS) workshop! 

This notebook covers the basics of using BRMS to fit Bayesian regression models. We'll build up from simple linear models to more complex hierarchical structures.

## [!] First Time Setup Notes

- **If prompted to install packages**: Your environment may need CmdStan. Reply **"y"** to install.
- **First model runtime**: ~30-60 seconds (compilation happens once)
- **Subsequent models**: <5 seconds (very fast!)
- **In container**: Everything pre-installed, should run without prompts
- **In Binder**: May prompt for first-time CmdStan setup (normal)

---

## Environment Check

If you encounter any installation prompts, run this chunk to diagnose:

```{r diagnose_environment, eval=FALSE}
# Uncomment and run if you see installation prompts

cat("=== Environment Diagnostics ===\n\n")

cat("R Version:\n")
print(R.version$version.string)

cat("\nCmdStan Installation Status:\n")
if (requireNamespace("cmdstanr", quietly = TRUE)) {
  cmdstanr_path <- try(cmdstanr::cmdstan_path(), silent = TRUE)
  if (inherits(cmdstanr_path, "try-error")) {
    cat("- cmdstanr installed but CmdStan not found\n")
    cat("- Run: cmdstanr::install_cmdstan()\n")
  } else {
    cat("✓ CmdStan found at:", cmdstanr_path, "\n")
  }
} else {
  cat("- cmdstanr not installed\n")
  cat("- Run: install.packages('cmdstanr')\n")
}

cat("\nBRMS Configuration:\n")
cat("- Backend:", getOption("brms.backend", "default"), "\n")
cat("- Cores:", getOption("mc.cores", "default"), "\n")
```

---

# Part 1: Data Preparation

First, let's load required packages and prepare our data.

```{r load_packages, message=TRUE}
cat("Loading required packages...\n")

# Load packages with better error handling
tryCatch({
  library(brms)        # Bayesian Regression Models using Stan
  library(tidyverse)   # Data manipulation
  library(bayesplot)   # Posterior visualization
  library(ggplot2)     # Plotting
  cat("✓ All packages loaded successfully\n")
}, error = function(e) {
  cat("Error loading packages:\n")
  print(e)
  cat("\nIf prompted to install missing packages, reply 'y' to install.\n")
})
```

Let's create a simple dataset to work with:

```{r create_data}
# Simulated data: effect of study time on exam scores
set.seed(42)
n <- 100

data_study <- tibble(
  student_id = 1:n,
  study_hours = rnorm(n, mean = 5, sd = 2),
  score = 40 + 8 * study_hours + rnorm(n, 0, 10)
) %>%
  mutate(
    study_hours = pmax(study_hours, 0),  # Can't have negative study hours
    score = pmin(pmax(score, 0), 100)     # Scores between 0-100
  )

# Explore the data
head(data_study, 10)
summary(data_study)
```

```{r plot_data}
ggplot(data_study, aes(x = study_hours, y = score)) +
  geom_point(size = 3, alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Relationship between Study Time and Exam Score",
    x = "Hours Studied",
    y = "Exam Score"
  )
```

---

# Part 2: Simple Linear Regression

Let's fit a Bayesian linear regression model. This is our first BRMS model!

**Note**: The first model takes ~30-60 seconds to compile, then subsequent models are fast.

```{r fit_model1, cache=TRUE}
cat("Fitting Bayesian linear regression...\n")

model1 <- brm(
  formula = score ~ study_hours,
  data = data_study,
  family = gaussian(),
  chains = 2,      # 2 parallel chains for speed (normal: 4)
  iter = 2000,     # Total iterations per chain
  warmup = 1000,   # Burn-in iterations
  cores = 2,       # Parallel processing
  refresh = 0      # Suppress iteration printing
)

cat("Model fitting complete!\n")
```

Let's examine the model:

```{r examine_model1}
# Summary of posterior distributions
summary(model1)
```

---

# Part 3: Posterior Visualization

Visualize the posterior distributions:

```{r posterior_plot1}
# Plot posterior distributions
plot(model1)
```

Extract and visualize specific parameters:

```{r posterior_extract}
# Extract posterior samples
posterior_samples <- as_draws_df(model1)
head(posterior_samples)

# Plot intercept and slope
posterior_samples %>%
  ggplot(aes(x = b_study_hours)) +
  geom_density(fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Posterior Distribution of Study Hours Effect",
    x = "Effect Size (slope)",
    y = "Density"
  )
```

---

# Part 4: Custom Priors

Specify custom priors for the model:

```{r check_priors}
# Check default priors
prior_summary(model1)

# Define custom priors
custom_priors <- c(
  prior(normal(50, 10), class = "Intercept"),
  prior(normal(8, 2), class = "b"),
  prior(exponential(1), class = "sigma")
)

# Model with custom priors
model2 <- brm(
  formula = score ~ study_hours,
  data = data_study,
  family = gaussian(),
  prior = custom_priors,
  chains = 2,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  refresh = 0
)

cat("Model 2 with custom priors complete!\n")
```

---

# Part 5: Model Comparison

Compare models using Leave-One-Out (LOO) cross-validation:

```{r compare_models}
# Add LOO criterion
model1_loo <- add_criterion(model1, "loo")
model2_loo <- add_criterion(model2, "loo")

# Compare
loo_compare(model1_loo, model2_loo)
```

---

# Part 6: Predictions

Make predictions with the model:

```{r predictions}
# Create new data for prediction
new_data <- data.frame(
  study_hours = seq(0, 12, by = 2)
)

# Get predicted values with uncertainty
predictions <- posterior_predict(model1, newdata = new_data, re_formula = NA)

# Calculate summaries
pred_summary <- data.frame(
  study_hours = new_data$study_hours,
  mean = colMeans(predictions),
  lower = apply(predictions, 2, quantile, probs = 0.025),
  upper = apply(predictions, 2, quantile, probs = 0.975)
)

# Plot predictions
ggplot(pred_summary, aes(x = study_hours, y = mean)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "steelblue") +
  geom_point(data = data_study, aes(x = study_hours, y = score), alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Bayesian Predictions with 95% Credible Interval",
    x = "Hours Studied",
    y = "Predicted Score"
  )
```

---

# Summary

Congratulations! You've learned:

1. ✅ Loading data and fitting a Bayesian regression model
2. ✅ Understanding posterior distributions
3. ✅ Visualizing results
4. ✅ Specifying custom priors
5. ✅ Comparing models with LOO
6. ✅ Making predictions with uncertainty

## Next Steps

- Explore hierarchical models for grouped data
- Fit models with different families (Poisson, binomial, etc.)
- Use regularizing priors for complex models
- Check model diagnostics (Rhat, ESS)

## Resources

- **BRMS Documentation**: https://paul-buerkner.github.io/brms/
- **Stan Manual**: https://mc-stan.org/docs/
- **Bayesian Visualization**: https://mc-stan.org/bayesplot/articles/

---

**Happy Bayesian modeling!** 🎓
