---
title: "Session 29: lme4 Model Criticism & Advanced Mixed-Effects Diagnostics"
subtitle: "Research Data and Methods Workshop Series — Collaborative Research Centre (CRC 1252) 'Prominence in Language'"
author: "Job Schepens"
date: "2026-09-30"
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-summary: "Toggle Code"
    theme: flatly
    embed-resources: true
    link-external-newwindow: true
execute:
  warning: false
  message: false
  cache: true
---

# Introduction: Model Criticism in Quantitative Linguistics

Linear Mixed-Effects Models (LMMs) and Generalized Linear Mixed-Effects Models (GLMMs) fitted via `lme4` [@bates2015fitting] are the cornerstone of contemporary quantitative and experimental linguistics. They empower researchers to model hierarchically clustered data, handle unbalanced designs, and simultaneously account for variation across speakers and linguistic materials.

However, standard summary tables can easily mask fatal assumption violations. A model that converges without error may violate core Gauss-Markov assumptions, suffer from degenerate random-effects geometry (singularity), or be driven by a handful of influential participants or idiosyncratic stimulus items [@meteyard2020best].

::: {.callout-note}
## Workshop Datasets: Authentic Linguistic Foundations for CRC 1252
Rather than generic non-linguistic toy examples (such as the `sleepstudy`, which lacks item variation and linguistic structure), this workshop demonstrates all diagnostic workflows on three authentic linguistic datasets:

1. **Continuous Reaction Time LMM (`languageR::lexdec`)** [@baayen2006morphological]: Visual lexical decision reaction times ($N = 1,659$) with **crossed random factors** (21 `Subject`s crossed with 79 `Word`s), continuous linguistic predictors (Lexical Frequency, Word Length), participant background (`NativeLanguage`), and experimental progression (`Trial`).
2. **Discrete Syntactic Choice GLMM (`languageR::dative`)** [@bresnan2007predicting]: Dative alternation choices ($N = 3,263$; double object NP *give John the book* vs. prepositional PP *give the book to John*), directly testing **prominence hierarchies** central to CRC 1252 (Animacy, Pronominality, Givenness, End-Weight).
3. **Ordinal Subjective Rating CLMM (`languageR::sizeRatings`)**: Semantic size ratings ($N = 3,078$) on a 7-point Likert scale across 38 subjects and 81 concrete nouns, demonstrating why naive Gaussian models fail on bounded ordinal responses.
:::

---

# Methodological Landscape: A Decade of Mixed-Effects Developments (2014–2024+)

Over the past decade, mixed-effects modeling in experimental linguistics and psychology has undergone profound methodological advancements. Understanding these developments clarifies why contemporary model criticism goes far beyond legacy residual checks:

## 1. The Random-Effects Complexity Debate (2013–2017)
In a landmark paper, @barr2013random argued that confirmatory hypothesis testing requires "maximal" random-effects structures—including random intercepts and random slopes for all within-unit factors—to avoid Type I error rates soaring to 20–40%. However, @bates2015parsimonious and @matuschek2017balancing demonstrated through extensive simulations that fitting maximal models on finite empirical datasets frequently produces **overparameterization**, severe loss of statistical power, and **singular fits**. They advocated for principled, data-driven model selection to strike an optimal balance between Type I error control and power.

## 2. The GLMM Residual Revolution: DHARMa (2016–2022)
For decades, researchers struggled to diagnose Generalized Linear Mixed Models (GLMMs) because raw, Pearson, and deviance residuals fail catastrophically on discrete binary and count outcomes due to intrinsic mean-variance coupling. Florian Hartig [-@hartig2022dharma] solved this by introducing simulation-based quantile residuals using the **Probability Integral Transform (PIT)**. This transformed GLMM diagnostics from an intractable guessing game into a rigorous testing suite for uniformity, overdispersion, and zero-inflation.

## 3. Ben Bolker’s Convergence Tooling & glmmTMB
As lead maintainer of `lme4`, Ben Bolker pioneered diagnostic tooling to demystify optimization warnings [@bolker2009generalized]. Key developments include:
- `lme4::allFit()`: Benchmarking models across multiple derivative-free optimizers to detect false-positive convergence warnings.
- `lme4::isSingular()`: Mathematical boundary checking based on Cholesky factor singular values.
- `glmmTMB` [@brooks2017glmmtmb]: Flexible template-model-builder architecture handling complex dispersion, zero-inflation, and spatial/temporal autocorrelation structures.
- `trouBBlme4SolveR`: Automated diagnostic routines to resolve boundary singularities.

## 4. Unified Model Criticism via `easystats` (2020–2024)
The `easystats` ecosystem [@ludecke2021performance] unified disparate diagnostic scripts into coherent workflows:
- `performance::check_model()`: A comprehensive visual audit covering linearity, homoscedasticity, posterior predictive checks, collinearity, and random-effect normality.
- `performance::r2_nakagawa()`: Formalized variance partitioning into Marginal and Conditional $R^2$ [@nakagawa2013general].

## 5. Post-Fitting & Marginal Effects (`emmeans` & `marginaleffects`)
Russell Lenth’s `emmeans` [-@lenth2016least] and Vincent Arel-Bundock’s `marginaleffects` [-@arelbundock2024marginaleffects] established modern protocols for post-hoc contrasts, reference grids, and counterfactual prediction on the response scale.

## 6. The Bayesian Migration: `brms` (2017–2024)
Paul-Christian Bürkner’s `brms` [-@burkner2017brms] opened full Hamiltonian Monte Carlo (HMC) estimation via Stan to applied linguists. As @vasishth2018bayesian emphasize, weakly informative priors naturally regularize boundary estimates, eliminating frequentist singular fits without requiring researchers to discard scientifically motivated random slopes.

---

# Comparative Assumptions Matrix: ANOVA vs. lme4 vs. brms

The table below summarizes how assumptions have evolved from classical ANOVA to modern Bayesian mixed models:

| Assumption Dimension | RM-ANOVA | `lme4` Linear (LMM) | `lme4` Generalized (GLMM) | Bayesian Mixed Models (`brms`) | Modern Diagnostic Tool |
| :--- | :--- | :--- | :--- | :--- | :--- |
| **Design Balance** | Strict balance required; listwise deletion for missing cells | Unbalanced cells and missing observations handled natively | Unbalanced cells and missing observations handled natively | Unbalanced cells and missing observations handled natively | Missing data audits (`naniar`, `summary`) |
| **Sphericity** | Required; violated sphericity inflates $F$-test Type I errors | **Relaxed**: Variance-covariance components modeled directly | **Relaxed**: Variance-covariance components modeled directly | **Relaxed**: Full covariance prior (e.g. LKJ) modeled directly | Mauchly's test (ANOVA only; bypassed in mixed models) |
| **Crossed Item Variation** | **Violated**: Language-as-fixed-effect fallacy [@clark1973language] | **Satisfied**: Crossed random intercepts & slopes (`Subject` + `Word`) | **Satisfied**: Crossed random structures | **Satisfied**: Crossed random structures with regularizing priors | Crossed formula syntax: `(1 \| Subject) + (1 \| Word)` |
| **Residual Distribution** | $\epsilon \sim \mathcal{N}(0, \sigma^2)$ | $\epsilon \sim \mathcal{N}(0, \sigma^2)$ (Normality required) | Uniform quantile residuals $\mathcal{U}(0,1)$ via PIT | Full posterior predictive distribution matching data family | Q-Q plots, `performance::check_model()`, `DHARMa::testUniformity()` |
| **Homoscedasticity** | Homogeneity across groups required | Homoscedastic residual variance $\sigma^2$ assumed | Variance coupled to mean ($\mu(1-\mu)$); overdispersion fatal | Can model residual variance as distributional regression (`sigma ~ ...`) | Scale-location plots, `DHARMa::testDispersion()` |
| **Linearity & Scaling** | Linear relationships assumed | Linear on response scale; scale continuous predictors | Linear on **link scale** (log, logit, probit, cloglog) | Linear or non-linear on link scale | Partial residual plots, `check_model()` |
| **Random Effects Normality** | N/A | BLUPs assumed multivariate normal: $b \sim \mathcal{N}(0, \Sigma)$ | Conditional modes assumed multivariate normal | Priors specify random-effects distributions (Normal, Student-$t$) | Q-Q plots of BLUPs, `check_model()` |
| **Singularity & Boundaries** | N/A | **Critical vulnerability**: Boundary collapse ($\sigma_b^2 = 0, \rho = \pm 1$) | **Critical vulnerability**: Laplace boundary failure | **Immune**: Regularizing priors keep estimates off hard boundary | `lme4::isSingular()`, `trouBBlme4SolveR::dwmw()` |
| **Residual Independence** | Assumed across trials | Assumed; sequential trial fatigue violates independence | Assumed; temporal/spatial clustering violates independence | Autoregressive structures can be added (`ar()`, `cor_ar()`) | `acf()`, `DHARMa::testTemporalAutocorrelation()` |

---

# Environment Setup

```{r setup, message=FALSE, warning=FALSE}
# Core mixed-effects modeling
library(lme4)
library(lmerTest)

# Simulation-based residual diagnostics
library(DHARMa)

# Multi-panel model checks & Nakagawa R-squared
library(performance)
library(see)

# Cluster-level influence diagnostics
library(influence.ME)

# Automated singularity & convergence troubleshooting
library(trouBBlme4SolveR)
library(dfoptim)

# Ordinal mixed models
library(ordinal)

# Post-hoc contrasts and estimated marginal means
library(emmeans)

# Linguistic benchmark datasets
library(languageR)

# Tidyverse data wrangling and plotting
library(tidyverse)

# Reproducibility seed & visualization theme
set.seed(12345)
theme_set(theme_minimal(base_size = 12))
```

---

# Module 1: Continuous Reaction Times & The Singularity Debate

## 1.1 Preparing Linguistic Lexical Decision Data
## 1.1 Preparing Linguistic Lexical Decision Data & The Need for Contrast Coding

We prepare `languageR::lexdec` [@baayen2006morphological], standardizing continuous predictors to optimize numerical conditioning and applying centered sum coding (-0.5 / +0.5) to categorical factors:
We prepare `languageR::lexdec` [@baayen2006morphological], standardizing continuous predictors to optimize numerical conditioning and applying **centered sum contrast coding** ($-0.5 / +0.5$) to categorical factors [@schad2020contrast]:

```{r prep-lexdec}
data(lexdec)

lexdec <- lexdec %>%
  mutate(
    Freq_z   = as.numeric(scale(Frequency)),
    Length_z = as.numeric(scale(Length)),
    Native_num = ifelse(NativeLanguage == "Other", 0.5, -0.5),
    Sex_num    = ifelse(Sex == "M", 0.5, -0.5)
  )

glimpse(select(lexdec, Subject, Word, RT, Trial, Freq_z, Length_z, Native_num, Sex_num))
```

::: {.callout-important}
## Why Contrast Coding is Essential Before Model Fitting
In R, factors default to **treatment (dummy) coding** (`contr.treatment`), which maps the first alphabetical level to 0 and all other levels to 1. In mixed-effects modeling, this default creates three major hazards [@schad2020contrast]:

1. **Misinterpretation of Main Effects**: In models with interactions ($A \times B$), the lower-order $\beta$ coefficients under treatment coding are **not** main effects; they are *simple effects* restricted specifically to the arbitrary baseline level of the other factor.
2. **Artificial Random Slope Singularity**: In `(1 + Condition | Subject)`, treatment coding forces the random intercept to represent variance *specifically at the baseline condition*. This induces severe collinearity between the random intercept and random slope, driving the estimated correlation parameter $\rho$ to boundary values of $\pm 1.0$ and triggering false singular fits.
3. **Invalid Type III ANOVA Tables**: Type III tests (`car::Anova(type = "III")`) mathematically assume orthogonal contrasts (`contr.sum`). Under default treatment coding, Type III tests are statistically meaningless.

Centered sum coding ($-0.5 / +0.5$) centers the intercept at the unweighted grand mean across conditions, decouples random intercepts from random slopes, and ensures $\beta$ coefficients represent true main effects.
:::

## 1.2 Fitting a Maximal Model (Barr et al., 2013)

According to @barr2013random, any factor varying within a grouping unit must have a corresponding random slope:
- Within `Subject`: `Freq_z`, `Length_z`
- Within `Word`: `Native_num`, `Sex_num`

```{r fit-maximal}
m_maximal <- lmer(
  RT ~ Freq_z + Length_z + Native_num + Sex_num +
    (1 + Freq_z + Length_z | Subject) +
    (1 + Native_num + Sex_num | Word),
  data = lexdec
)

# Test for singularity
cat("Is the maximal model singular?", isSingular(m_maximal), "\n")
```

## 1.3 Inspecting Boundary Singularity

When `isSingular()` returns `TRUE`, the variance-covariance matrix lies on the boundary:

```{r inspect-varcorr}
print(VarCorr(m_maximal))
```

::: {.callout-warning}
## Diagnostic Finding: Singularity Collapse
Notice that the `Word` random intercept standard deviation has collapsed to `0.000` with a boundary correlation of `1.000`. The dataset cannot support estimating independent item-level slopes for speaker background factors alongside intercepts [@bates2015parsimonious].
:::

## 1.4 Automated Singularity Resolution (`trouBBlme4SolveR`)

The `trouBBlme4SolveR::dwmw()` function diagnoses which random slope has zero variance and systematically prunes it:

```{r dwmw-resolution}
m_resolved <- dwmw(m_maximal)
cat("\nIs resolved model singular?", isSingular(m_resolved), "\n")
print(summary(m_resolved))
```

---

# Module 2: Simulation Diagnostics with DHARMa

## 2.1 Classical Residual Checks in Linear Mixed Models

In standard Gaussian LMMs, residuals are assumed to follow $\epsilon \sim \mathcal{N}(0, \sigma^2)$:

```{r fit-parsimonious}
m_parsimonious <- lmer(
  RT ~ Freq_z + Length_z + Native_num + (1 + Freq_z | Subject) + (1 | Word),
  data = lexdec
)
```

```{r classical-residual-plots, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
qqnorm(residuals(m_parsimonious), main = "Normal Q-Q Plot (Raw Residuals)")
qqline(residuals(m_parsimonious), col = "red", lwd = 2)

plot(fitted(m_parsimonious), residuals(m_parsimonious),
     xlab = "Fitted Values (log RT)", ylab = "Residuals",
     main = "Residuals vs. Fitted", pch = 16, col = rgb(0,0,0,0.2))
abline(h = 0, col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
```

## 2.2 Discrete GLMM Residual Crisis & Probability Integral Transform

In discrete GLMMs (e.g. logistic regression), raw and Pearson residuals form parallel diagonal bands because outcomes are binary ($y \in \{0, 1\}$).

The **DHARMa** package [@hartig2022dharma] uses the **Probability Integral Transform (PIT)** to simulate $n$ datasets from the fitted model and compute quantile residuals:
$$u_i = P(Y^{\text{sim}}_i \le y_i) \sim \mathcal{U}(0, 1)$$

## 2.3 DHARMa in Action: Prominence in the Dative Alternation

We model the English dative alternation [@bresnan2007predicting], where prominence features (animate recipients, pronominal recipients, short themes) predict the double-object construction (NP) versus prepositional construction (PP):

```{r fit-glmm-dative}
data(dative)

m_dative <- glmer(
  RealizationOfRecipient ~ AnimacyOfRec + PronomOfRec + LengthOfTheme + (1 | Verb),
  family = binomial,
  data = dative
)
summary(m_dative)
```

```{r dharma-sim-glmm, fig.width=10, fig.height=5}
sim_dative <- simulateResiduals(fittedModel = m_dative, n = 500, plot = FALSE)
plot(sim_dative)
```

```{r dharma-formal-tests}
# Test residual uniformity (Kolmogorov-Smirnov)
testUniformity(sim_dative)

# Test overdispersion / underdispersion
testDispersion(sim_dative)

# Test excess zeros
testZeroInflation(sim_dative)
```

## 2.4 Testing Sequential Autocorrelation across Trials

In self-paced reading and reaction time experiments, participant fatigue often induces temporal autocorrelation. We aggregate residuals across participants for each experimental trial:

```{r dharma-temporal-autocorr, fig.width=9, fig.height=4}
sim_lexdec <- simulateResiduals(fittedModel = m_parsimonious, n = 250, plot = FALSE)
sim_trial <- recalculateResiduals(sim_lexdec, group = lexdec$Trial)
testTemporalAutocorrelation(sim_trial, time = unique(lexdec$Trial))

# Individual-level autocorrelation (ACF) for participant A1
res_a1 <- residuals(m_parsimonious)[lexdec$Subject == "A1"]
acf(res_a1, main = "Residual Autocorrelation (Subject A1)")
```

---

# Module 3: Multicollinearity, Multi-Panel Checks & Cluster Influence

## 3.1 Multicollinearity & Variance Inflation Factors (VIF)

In natural language, lexical features naturally correlate (Zipf's Law: frequent words are shorter). We test Variance Inflation Factors using `performance::check_collinearity()`:

```{r check-collinearity}
vif_results <- check_collinearity(m_parsimonious)
print(vif_results)
plot(vif_results)
```

## 3.2 Unified Multi-Panel Diagnostics (`easystats`)

```{r multi-panel-check, fig.width=10, fig.height=9}
check_model(m_parsimonious)
```

## 3.3 Cluster-Level Influence & Cook's Distance (`influence.ME`)

We check whether particular speakers or items exert disproportionate leverage [@nieuwenhuis2012influence]:

```{r influence-subjects, fig.width=9, fig.height=4.5}
infl_subj <- influence(m_parsimonious, group = "Subject", count = FALSE)
plot(infl_subj, which = "cook", cutoff = 4 / length(unique(lexdec$Subject)),
     main = "Subject-Level Cook's Distance (lexdec)")
```

```{r influence-words, fig.width=9, fig.height=4.5}
infl_word <- influence(m_parsimonious, group = "Word", count = FALSE)
plot(infl_word, which = "cook", cutoff = 4 / length(unique(lexdec$Word)),
     main = "Word-Level Cook's Distance (Items)")
```

---

# Module 4: Ordinal Responses & Cumulative Link Models

## 4.1 The Likert Assumption Fallacy

Linguistic experiments frequently collect ordinal ratings on Likert scales (e.g. acceptability, semantic size, prototypicality). A widespread practice is treating 1–7 Likert ratings as metric Gaussian variables via `lme4::lmer()`.

This practice violates three fundamental assumptions:
1. **Bounded Support**: Data are bounded at 1 and 7 (floor/ceiling distortions).
2. **Non-Interval Metric**: The cognitive distance between "1" and "2" is not guaranteed to equal the distance between "4" and "5".
3. **Discrete Latent Thresholds**: Responses reflect an underlying continuous latent variable $Y^*$ partitioned by cutpoints $\theta_1 < \theta_2 < \dots < \theta_{K-1}$.

## 4.2 Demonstrating the Flaw: Gaussian LMM on Likert Data

Using `languageR::sizeRatings` (3,078 size ratings of 81 words across 38 subjects on a 1–7 scale):

```{r naive-gaussian-ordinal}
data(sizeRatings)

# Naive Gaussian LMM
m_naive_likert <- lmer(Rating ~ Class + (1 | Word), data = sizeRatings)

# DHARMa diagnostic plot on naive Gaussian fit
sim_naive <- simulateResiduals(m_naive_likert, n = 250, plot = FALSE)
plot(sim_naive)
```

::: {.callout-caution}
## Diagnostic Warning: Severe Quantile Distortion
The DHARMa Q-Q plot on the naive Gaussian fit exhibits severe, statistically significant S-shaped curvature and failed uniformity tests ($p < 10^{-16}$), proving that standard Gaussian mixed models fail on bounded Likert data.
:::

## 4.3 The Principled Solution: Cumulative Link Mixed Models (`ordinal::clmm`)

We fit a proper Cumulative Link Mixed Model using `ordinal::clmm` [@christensen2019ordinal]:

$$\text{logit}(P(Y_i \le k)) = \theta_k - (X_i \beta + Z_i u)$$

```{r clmm-fit}
sizeRatings <- sizeRatings %>%
  mutate(Rating_ord = factor(Rating, ordered = TRUE))

m_clmm <- clmm(Rating_ord ~ Class + (1 | Word), data = sizeRatings)
summary(m_clmm)
```

The model estimates the fixed effect of semantic class alongside the 6 latent cutpoints ($\theta_{1|2}$ through $\theta_{6|7}$), properly preserving ordinal geometry.

---

# Module 5: Post-Fitting Inference with emmeans
# Module 5: Post-Fitting Inference with emmeans & The Contrast Coding Synthesis

## 5.1 Estimating Marginal Means & Contrasts

The `emmeans` package [@lenth2016least] computes Estimated Marginal Means (least-squares means) by averaging predictions over a balanced reference grid:

```{r emmeans-lmm}
# Pairwise comparisons on continuous RT model
emm_native <- emmeans(m_parsimonious, ~ Native_num)
print(emm_native)
pairs(emm_native)
```

For Generalized and Ordinal models, we compute predictions on the latent link scale or back-transform to the response scale:

```{r emmeans-glmm}
# Dative realization: back-transformed response probabilities
emm_dative <- emmeans(m_dative, ~ AnimacyOfRec, type = "response")
print(emm_dative)
pairs(emm_dative)
```

```{r emmeans-clmm}
# Ordinal model: latent linear predictor contrasts
emm_clmm <- emmeans(m_clmm, ~ Class)
print(emm_clmm)
pairs(emm_clmm)
```

## 5.2 Methodological Assumptions of emmeans
## 5.2 Empirical Demonstration: Coding Invariance of emmeans

A profound property of `emmeans` is that it evaluates predictions over a balanced reference grid regardless of how factors were coded in the design matrix $\mathbf{X}$. 

Let us compare two identical models fitted with default treatment coding vs. centered sum coding:

```{r contrast-coding-vs-emmeans}
# 1. Model with default treatment coding (English = 0, Other = 1)
m_treatment <- lmer(RT ~ Freq_z * NativeLanguage + (1 | Subject), data = lexdec)

# 2. Model with centered sum coding (-0.5 / +0.5)
m_sum <- lmer(RT ~ Freq_z * Native_num + (1 | Subject), data = lexdec)

# In the model tables, the lower-order coefficients differ completely:
# In m_treatment: 'NativeLanguageOther' is a simple effect at Freq_z = 0
# In m_sum: 'Native_num' is a true main effect averaged across Freq_z
fixef(m_treatment)
fixef(m_sum)

# Yet emmeans produces the EXACT SAME estimated marginal means:
emm_from_treat <- emmeans(m_treatment, ~ NativeLanguage)
emm_from_sum   <- emmeans(m_sum, ~ Native_num)

cat("\nemmeans from Treatment-Coded Model:\n")
print(emm_from_treat)

cat("\nemmeans from Sum-Coded Model:\n")
print(emm_from_sum)
```

## 5.3 Methodological Assumptions of emmeans

When reporting marginal means, researchers must verify five underlying assumptions:

1. **Reference Grid Balancing**: Continuous covariates are held constant at their grand mean ($\bar{x}$), and factor cells are weighted equally. When covariates are correlated, the reference grid can evaluate combinations unobserved in actual data.
2. **Scale of Inference**: Differences on the link scale (log-odds) back-transform to **odds ratios** or **ratio of medians** on the response scale.
3. **Degrees of Freedom Approximations**: For LMMs, `emmeans` supports **Kenward-Roger** and **Satterthwaite** approximations [@kuznetsova2017lmertest]. Kenward-Roger is preferred in unbalanced crossed designs to avoid anticonservative tests.
4. **Homoscedasticity & Pooled Variance**: Standard errors in `emmeans` assume a shared residual variance $\sigma^2$ across factor levels.
5. **Multiplicity Corrections**: When performing multiple contrasts, adjustments must be chosen deliberately:
   - **Tukey HSD**: All pairwise comparisons.
   - **Bonferroni / Šidák**: Planned orthogonal contrasts.
   - **Benjamini-Hochberg (FDR)**: Large exploratory contrast families.

## 5.4 Why Both Are Necessary: The Two-Stage Principle

If `emmeans` produces the same marginal means regardless of factor coding, **why do we still need contrast coding before fitting?** Conversely, if we use sum contrast coding, **why do we still need `emmeans`?**

The answer lies in the distinction between **model estimation** (*a priori*) and **hypothesis evaluation** (*post hoc*):

| Dimension / Objective | Contrast Coding (Before Fitting) | `emmeans` (After Fitting) | Why Both Are Essential |
| :--- | :--- | :--- | :--- |
| **Operational Stage** | *A priori*: Parameterizes the design matrix $\mathbf{X}$ during numerical optimization. | *Post hoc*: Constructs reference grid predictions from estimated $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{\Sigma}}$. | Neither tool can perform the other's job: `emmeans` cannot refit the model; contrast coding cannot predict across balanced grids automatically. |
| **Random Slope Stability** | **Crucial**: Centered coding centers random intercepts at the grand mean, preventing artificial correlation collapse ($\rho \approx \pm 1.0$) and singular fits. | **Cannot help**: Operates only on the already-converged model. If uncentered coding triggered a singular fit, `emmeans` cannot repair the corrupted variance parameters. | You **must** contrast-code before fitting to ensure the model converges to valid, non-singular variance estimates. |
| **Coefficient Interpretation** | Ensures lower-order $\beta$ terms in `summary()` represent true main effects rather than simple effects at an arbitrary baseline. | Does not alter the $\beta$ regression table; only provides reconstructed marginal predictions. | Researchers reading model tables require interpretable $\beta$ coefficients for regression reporting. |
| **Type III ANOVA (`car::Anova`)** | **Strictly Required**: Type III tests mathematically assume sum/orthogonal contrasts (`contr.sum`). Under treatment coding, Type III tests are statistically meaningless. | Not applicable (evaluates specific linear contrasts rather than overall omnibus decomposition). | If journal reviewers demand Type III Wald/LRT ANOVA tables, models must be fitted with sum contrasts. |
| **Interaction Decomposition** | Slicing interactions via contrast matrices requires tedious manual coding of custom linear combinations ($\mathbf{c}^\top \boldsymbol{\beta}$) or repeatedly releveling baselines. | **Trivial**: Syntax like `pairwise ~ FactorA \| FactorB` decomposes multi-way interactions into simple pairwise differences in a single line. | You need `emmeans` to explore interactions cleanly without refitting the model multiple times. |
| **Response Scale for GLMMs** | Operates strictly on the link scale (log-odds $\eta$). Back-transforming manually requires tedious delta-method calculations for SEs. | **Native**: Converts log-odds to response probabilities ($p \in [0, 1]$) and odds ratios with automatic multivariate delta-method standard errors. | You need `emmeans` to translate abstract log-odds coefficients into meaningful probabilities for publication. |
| **Multiplicity & d.f. Corrections** | Manual adjustment of $p$-values and manual calculation of Satterthwaite/Kenward-Roger d.f. | Automated integration with `lmerTest`/`pbkrtest` for Satterthwaite/Kenward-Roger d.f. and Tukey/FDR adjustments. | You need `emmeans` to avoid anticonservative inferential tests. |

::: {.callout-tip}
## The Gold-Standard Two-Stage Workflow
1. **Stage 1 (Before Model Fitting):** Apply centered sum contrast coding (`contr.sum` or $-0.5 / +0.5$) to all categorical predictors. This stabilizes the random-effects covariance matrix against boundary singularities, aligns $\beta$ coefficients with main effects, and validates Type III ANOVA tables.
2. **Stage 2 (After Model Fitting):** Use `emmeans` to evaluate condition means across balanced reference grids, unpack complex interactions, compute response-scale probabilities and odds ratios, and apply Kenward-Roger d.f. approximations and Tukey multiplicity adjustments.
:::

---

# Module 6: Optimizer Benchmarking, Variance Partitioning & Reporting

## 6.1 Optimizer Benchmarking with `allFit()`

```{r benchmark-allfit, message=FALSE, warning=FALSE}
all_fits <- allFit(m_parsimonious)
summary(all_fits)
```

## 6.2 Likelihood Ratio Tests: Enforcing REML vs. ML Rules

* **Testing Random Effects**: Keep fixed effects identical; use **`REML = TRUE`** with `anova(m1, m2)`.
* **Testing Fixed Effects**: Use **`REML = FALSE`** (Maximum Likelihood) with `anova(m1, m2)`.
* **Final Reporting**: Refit selected model with **`REML = TRUE`** for unbiased variance estimates.

```{r lrt-comparison}
m_full_ml <- lmer(RT ~ Freq_z + Length_z + Native_num + (1 | Subject) + (1 | Word),
                  data = lexdec, REML = FALSE)
m_null_ml <- lmer(RT ~ Length_z + Native_num + (1 | Subject) + (1 | Word),
                  data = lexdec, REML = FALSE)

anova(m_null_ml, m_full_ml)
```

## 6.3 Variance Partitioning: Nakagawa’s $R^2$

```{r nakagawa-r2}
r2_metrics <- r2_nakagawa(m_resolved)
print(r2_metrics)
```

## 6.4 Open Science Reporting Checklist [@meteyard2020best]

1. **Specification**: State full formula syntax, contrast coding schemes, and scaling transformations.
2. **Estimation**: Specify REML vs. ML, optimizer choice, and integration method.
3. **Diagnostics**: Report `isSingular()` checks, DHARMa dispersion/uniformity tests, and VIF metrics.
4. **Parameters**: Provide fixed effect estimates ($\beta$, $SE$, $t/z$, df approximation, $p$), random effect variance components ($\sigma, \rho$), and Nakagawa $R^2$ ($R^2_m, R^2_c$).
5. **Reproducibility**: Share annotated Quarto (`.qmd`) scripts and data in open repositories.

---

# References

::: {#refs}
:::
