---
title: "Session 29: lme4 Model Criticism & Advanced Mixed-Effects Diagnostics"
subtitle: "Research Data and Methods Workshop Series — Collaborative Research Centre (CRC 1252) 'Prominence in Language'"
author: "Job Schepens"
date: "2026-09-30"
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-summary: "Toggle Code"
    theme: flatly
    embed-resources: true
    link-external-newwindow: true
execute:
  warning: false
  message: false
  cache: true
---

# Introduction: Model Criticism in Quantitative Linguistics

Linear Mixed-Effects Models (LMMs) and Generalized Linear Mixed-Effects Models (GLMMs) fitted via `lme4` [@bates2015fitting] are the cornerstone of contemporary quantitative and experimental linguistics. They empower researchers to model hierarchically clustered data, handle unbalanced designs, and simultaneously account for variation across speakers and linguistic materials.

However, standard summary tables can easily mask fatal assumption violations. A model that converges without error may violate core Gauss-Markov assumptions, suffer from degenerate random-effects geometry (singularity), or be driven by a handful of influential participants or idiosyncratic stimulus items [@meteyard2020best].

::: {.callout-note}
## Workshop Datasets: Authentic Linguistic Foundations for CRC 1252
Rather than generic non-linguistic toy examples (such as the `sleepstudy`, which lacks item variation and linguistic structure), this workshop demonstrates all diagnostic workflows on three authentic linguistic datasets:

1. **Continuous Reaction Time LMM (`languageR::lexdec`)** [@baayen2006morphological]: Visual lexical decision reaction times ($N = 1,659$) with **crossed random factors** (21 `Subject`s crossed with 79 `Word`s), continuous linguistic predictors (Lexical Frequency, Word Length), participant background (`NativeLanguage`), and experimental progression (`Trial`).
2. **Discrete Syntactic Choice GLMM (`languageR::dative`)** [@bresnan2007predicting]: Dative alternation choices ($N = 3,263$; double object NP *give John the book* vs. prepositional PP *give the book to John*), directly testing **prominence hierarchies** central to CRC 1252 (Animacy, Pronominality, Givenness, End-Weight).
3. **Ordinal Subjective Rating CLMM (`languageR::sizeRatings`)**: Semantic size ratings ($N = 3,078$) on a 7-point Likert scale across 38 subjects and 81 concrete nouns, demonstrating why naive Gaussian models fail on bounded ordinal responses.
:::

---

# Methodological Landscape: A Decade of Mixed-Effects Developments (2014–2024+)

Over the past decade, mixed-effects modeling in experimental linguistics and psychology has undergone profound methodological advancements. Understanding these developments clarifies why contemporary model criticism goes far beyond legacy residual checks:

## 1. The Random-Effects Complexity Debate (2013–2017)
In a landmark paper, @barr2013random argued that confirmatory hypothesis testing requires "maximal" random-effects structures—including random intercepts and random slopes for all within-unit factors—to avoid Type I error rates soaring to 20–40%. However, @bates2015parsimonious and @matuschek2017balancing demonstrated through extensive simulations that fitting maximal models on finite empirical datasets frequently produces **overparameterization**, severe loss of statistical power, and **singular fits**. They advocated for principled, data-driven model selection to strike an optimal balance between Type I error control and power.

## 2. The GLMM Residual Revolution: DHARMa (2016–2022)
For decades, researchers struggled to diagnose Generalized Linear Mixed Models (GLMMs) because raw, Pearson, and deviance residuals fail catastrophically on discrete binary and count outcomes due to intrinsic mean-variance coupling. Florian Hartig [-@hartig2022dharma] solved this by introducing simulation-based quantile residuals using the **Probability Integral Transform (PIT)**. This transformed GLMM diagnostics from an intractable guessing game into a rigorous testing suite for uniformity, overdispersion, and zero-inflation.

## 3. Ben Bolker’s Convergence Tooling & glmmTMB
As lead maintainer of `lme4`, Ben Bolker pioneered diagnostic tooling to demystify optimization warnings [@bolker2009generalized]. Key developments include:
- `lme4::allFit()`: Benchmarking models across multiple derivative-free optimizers to detect false-positive convergence warnings.
- `lme4::isSingular()`: Mathematical boundary checking based on Cholesky factor singular values.
- `glmmTMB` [@brooks2017glmmtmb]: Flexible template-model-builder architecture handling complex dispersion, zero-inflation, and spatial/temporal autocorrelation structures.
- `trouBBlme4SolveR`: Automated diagnostic routines to resolve boundary singularities.

## 4. Unified Model Criticism via `easystats` (2020–2024)
The `easystats` ecosystem [@ludecke2021performance] unified disparate diagnostic scripts into coherent workflows:
- `performance::check_model()`: A comprehensive visual audit covering linearity, homoscedasticity, posterior predictive checks, collinearity, and random-effect normality.
- `performance::r2_nakagawa()`: Formalized variance partitioning into Marginal and Conditional $R^2$ [@nakagawa2013general].

## 5. Post-Fitting & Marginal Effects (`emmeans` & `marginaleffects`)
Russell Lenth’s `emmeans` [-@lenth2016least] and Vincent Arel-Bundock’s `marginaleffects` [-@arelbundock2024marginaleffects] established modern protocols for post-hoc contrasts, reference grids, and counterfactual prediction on the response scale.

## 6. The Bayesian Migration: `brms` (2017–2024)
Paul-Christian Bürkner’s `brms` [-@burkner2017brms] opened full Hamiltonian Monte Carlo (HMC) estimation via Stan to applied linguists. As @vasishth2018bayesian emphasize, weakly informative priors naturally regularize boundary estimates, eliminating frequentist singular fits without requiring researchers to discard scientifically motivated random slopes.

---

# Comparative Assumptions Matrix: ANOVA vs. lme4 vs. brms

The table below summarizes how assumptions have evolved from classical ANOVA to modern Bayesian mixed models:

| Assumption Dimension | RM-ANOVA | `lme4` Linear (LMM) | `lme4` Generalized (GLMM) | Bayesian Mixed Models (`brms`) | Modern Diagnostic Tool |
| :--- | :--- | :--- | :--- | :--- | :--- |
| **Design Balance** | Strict balance required; listwise deletion for missing cells | Unbalanced cells and missing observations handled natively | Unbalanced cells and missing observations handled natively | Unbalanced cells and missing observations handled natively | Missing data audits (`naniar`, `summary`) |
| **Sphericity** | Required; violated sphericity inflates $F$-test Type I errors | **Relaxed**: Variance-covariance components modeled directly | **Relaxed**: Variance-covariance components modeled directly | **Relaxed**: Full covariance prior (e.g. LKJ) modeled directly | Mauchly's test (ANOVA only; bypassed in mixed models) |
| **Crossed Item Variation** | **Violated**: Language-as-fixed-effect fallacy [@clark1973language] | **Satisfied**: Crossed random intercepts & slopes (`Subject` + `Word`) | **Satisfied**: Crossed random structures | **Satisfied**: Crossed random structures with regularizing priors | Crossed formula syntax: `(1 \| Subject) + (1 \| Word)` |
| **Residual Distribution** | $\epsilon \sim \mathcal{N}(0, \sigma^2)$ | $\epsilon \sim \mathcal{N}(0, \sigma^2)$ (Normality required) | Uniform quantile residuals $\mathcal{U}(0,1)$ via PIT | Full posterior predictive distribution matching data family | Q-Q plots, `performance::check_model()`, `DHARMa::testUniformity()` |
| **Homoscedasticity** | Homogeneity across groups required | Homoscedastic residual variance $\sigma^2$ assumed | Variance coupled to mean ($\mu(1-\mu)$); overdispersion fatal | Can model residual variance as distributional regression (`sigma ~ ...`) | Scale-location plots, `DHARMa::testDispersion()` |
| **Linearity & Scaling** | Linear relationships assumed | Linear on response scale; scale continuous predictors | Linear on **link scale** (log, logit, probit, cloglog) | Linear or non-linear on link scale | Partial residual plots, `check_model()` |
| **Random Effects Normality** | N/A | BLUPs assumed multivariate normal: $b \sim \mathcal{N}(0, \Sigma)$ | Conditional modes assumed multivariate normal | Priors specify random-effects distributions (Normal, Student-$t$) | Q-Q plots of BLUPs, `check_model()` |
| **Singularity & Boundaries** | N/A | **Critical vulnerability**: Boundary collapse ($\sigma_b^2 = 0, \rho = \pm 1$) | **Critical vulnerability**: Laplace boundary failure | **Immune**: Regularizing priors keep estimates off hard boundary | `lme4::isSingular()`, `trouBBlme4SolveR::dwmw()` |
| **Residual Independence** | Assumed across trials | Assumed; sequential trial fatigue violates independence | Assumed; temporal/spatial clustering violates independence | Autoregressive structures can be added (`ar()`, `cor_ar()`) | `acf()`, `DHARMa::testTemporalAutocorrelation()` |

---

# Environment Setup

```{r setup, message=FALSE, warning=FALSE}
# Core mixed-effects modeling
library(lme4)
library(lmerTest)

# Simulation-based residual diagnostics
library(DHARMa)

# Multi-panel model checks & Nakagawa R-squared
library(performance)
library(parameters)
library(see)

# Cluster-level influence diagnostics
library(influence.ME)

# Automated singularity & convergence troubleshooting
library(trouBBlme4SolveR)
library(dfoptim)

# Publication tables
library(knitr)

# Ordinal mixed models
library(ordinal)

# Post-hoc contrasts and estimated marginal means
library(emmeans)

# Linguistic benchmark datasets
library(languageR)

# Tidyverse data wrangling and plotting
library(tidyverse)

# Reproducibility seed & visualization theme
set.seed(12345)
theme_set(theme_minimal(base_size = 12))
```

---

# Module 1: Continuous Reaction Times & The Singularity Debate

## 1.1 Preparing Linguistic Lexical Decision Data & Random Effects Syntax

### Random Effects Syntax Reference
Mixed-effects models relax the rigid balanced-design and sphericity assumptions of repeated-measures ANOVA:
- **Nested random effects** (e.g., Students nested within Classrooms):
  `Formula: (1 | Class / Student)` or `(1 | Class) + (1 | Class:Student)`
- **Crossed random effects** (e.g., Participants crossed with Stimulus Words):
  `Formula: (1 | Subject) + (1 | Word)`

We prepare `languageR::lexdec` [@baayen2006morphological], standardizing continuous predictors to optimize numerical conditioning and applying **centered sum contrast coding** ($-0.5 / +0.5$) to categorical factors [@schad2020contrast]:

```{r prep-lexdec}
# Load lexical decision dataset from languageR (Baayen et al., 2006)
data(lexdec)

# 1. Rescale continuous predictors to z-scores (mean = 0, sd = 1)
#    This eliminates scale disparities and avoids ill-conditioned deviance valleys
# 2. Apply centered sum contrast coding (-0.5 / +0.5) to two-level factors
#    Ensures intercept represents grand mean and beta represents true main effects
lexdec <- lexdec %>%
  mutate(
    Freq_z     = as.numeric(scale(Frequency)),
    Length_z   = as.numeric(scale(Length)),
    Native_num = ifelse(NativeLanguage == "Other", 0.5, -0.5),
    Sex_num    = ifelse(Sex == "M", 0.5, -0.5)
  )

# Inspect processed variables
glimpse(select(lexdec, Subject, Word, RT, Trial, Freq_z, Length_z, Native_num, Sex_num))
```

::: {.callout-important}
## Why Contrast Coding is Essential Before Model Fitting
In R, factors default to **treatment (dummy) coding** (`contr.treatment`), which maps the first alphabetical level to 0 and all other levels to 1. In mixed-effects modeling, this default creates three major hazards [@schad2020contrast]:

1. **Misinterpretation of Main Effects**: In models with interactions ($A \times B$), the lower-order $\beta$ coefficients under treatment coding are **not** main effects; they are *simple effects* restricted specifically to the arbitrary baseline level of the other factor.
2. **Artificial Random Slope Singularity**: In `(1 + Condition | Subject)`, treatment coding forces the random intercept to represent variance *specifically at the baseline condition*. This induces severe collinearity between the random intercept and random slope, driving the estimated correlation parameter $\rho$ to boundary values of $\pm 1.0$ and triggering false singular fits.
3. **Invalid Type III ANOVA Tables**: Type III tests (`car::Anova(type = "III")`) mathematically assume orthogonal contrasts (`contr.sum`). Under default treatment coding, Type III tests are statistically meaningless.

Centered sum coding ($-0.5 / +0.5$) centers the intercept at the unweighted grand mean across conditions, decouples random intercepts from random slopes, and ensures $\beta$ coefficients represent true main effects.
:::

## 1.2 Fitting a Maximal Model (Barr et al., 2013)

According to @barr2013random, any factor varying within a grouping unit must have a corresponding random slope:
- Within `Subject`: `Freq_z`, `Length_z`
- Within `Word`: `Native_num`, `Sex_num`

```{r fit-maximal}
# Fit maximal random-effects model (Barr et al., 2013)
# Random intercepts and within-unit random slopes for all predictors:
# - Within Subject: Frequency and Word Length
# - Within Word: Native Language and Speaker Sex
m_maximal <- lmer(
  RT ~ Freq_z + Length_z + Native_num + Sex_num +
    (1 + Freq_z + Length_z | Subject) +
    (1 + Native_num + Sex_num | Word),
  data = lexdec
)

# Test for singularity (Cholesky factor on boundary)
cat("Is the maximal model singular?", isSingular(m_maximal), "\n")
```

## 1.3 Inspecting Boundary Singularity

When `isSingular()` returns `TRUE`, the variance-covariance matrix lies on the boundary:

```{r inspect-varcorr}
# Inspect estimated variance-covariance matrices and correlation parameters (rho)
# Look for variances collapsing to 0.000 or correlations at +/- 1.000
print(VarCorr(m_maximal))
```

::: {.callout-warning}
## Diagnostic Finding: Singularity Collapse
Notice that the `Word` random intercept standard deviation has collapsed to `0.000` with a boundary correlation of `1.000`. The dataset cannot support estimating independent item-level slopes for speaker background factors alongside intercepts [@bates2015parsimonious].
:::

## 1.4 Automated Singularity Resolution (`trouBBlme4SolveR`)

The `trouBBlme4SolveR::dwmw()` function diagnoses which random slope has zero variance and systematically prunes it:

```{r dwmw-resolution}
# Automated troubleshooting for singular fits (Ben Bolker's trouBBlme4SolveR)
# 'dwmw' ("Don't Worry, Make Warmer") iteratively tests and prunes zero-variance random slopes
m_resolved <- dwmw(m_maximal)
cat("\nIs resolved model singular?", isSingular(m_resolved), "\n")
print(summary(m_resolved))
```

---

# Module 2: Simulation Diagnostics with DHARMa

## 2.1 Classical Residual Checks in Linear Mixed Models

In standard Gaussian LMMs, residuals are assumed to follow $\epsilon \sim \mathcal{N}(0, \sigma^2)$:

```{r fit-parsimonious}
# Parsimonious random-effects model (Bates et al., 2015)
# Retains subject-level frequency slope; prunes overparameterized item slopes
m_parsimonious <- lmer(
  RT ~ Freq_z + Length_z + Native_num + (1 + Freq_z | Subject) + (1 | Word),
  data = lexdec
)
```

```{r classical-residual-plots, fig.width=9, fig.height=4}
# Classical diagnostic plots for Gaussian LMM:
# Left: Normal Q-Q plot to assess residual normality (epsilon ~ N(0, sigma^2))
# Right: Residuals vs. Fitted to assess homoscedasticity and linearity
par(mfrow = c(1, 2))
qqnorm(residuals(m_parsimonious), main = "Normal Q-Q Plot (Raw Residuals)")
qqline(residuals(m_parsimonious), col = "red", lwd = 2)

plot(fitted(m_parsimonious), residuals(m_parsimonious),
     xlab = "Fitted Values (log RT)", ylab = "Residuals",
     main = "Residuals vs. Fitted", pch = 16, col = rgb(0,0,0,0.2))
abline(h = 0, col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
```

## 2.2 Discrete GLMM Residual Crisis & Probability Integral Transform

In discrete GLMMs (e.g. logistic regression), raw and Pearson residuals form parallel diagonal bands because outcomes are binary ($y \in \{0, 1\}$).

The **DHARMa** package [@hartig2022dharma] uses the **Probability Integral Transform (PIT)** to simulate $n$ datasets from the fitted model and compute quantile residuals:
$$u_i = P(Y^{\text{sim}}_i \le y_i) \sim \mathcal{U}(0, 1)$$

## 2.3 DHARMa in Action: Prominence in the Dative Alternation

We model the English dative alternation [@bresnan2007predicting], where prominence features (animate recipients, pronominal recipients, short themes) predict the double-object construction (NP) versus prepositional construction (PP):

```{r fit-glmm-dative}
# Discrete Prominence GLMM: English Dative Alternation (Bresnan et al., 2007)
# Binary outcome: Realization of recipient as NP (double object) vs PP (prepositional)
data(dative)

m_dative <- glmer(
  RealizationOfRecipient ~ AnimacyOfRec + PronomOfRec + LengthOfTheme + (1 | Verb),
  family = binomial,
  data = dative
)
summary(m_dative)
```

```{r dharma-sim-glmm, fig.width=10, fig.height=5}
# DHARMa: Simulate standardized quantile residuals via Probability Integral Transform (PIT)
# Residuals are transformed to Uniform U(0, 1) under the null of correct specification
sim_dative <- simulateResiduals(fittedModel = m_dative, n = 500, plot = FALSE)

# DHARMa 2-panel diagnostic plot: Uniform Q-Q plot and Residuals vs. Predicted with spline
plot(sim_dative)
```

```{r dharma-formal-tests}
# 1. Test overall uniformity (Kolmogorov-Smirnov test against U(0, 1))
testUniformity(sim_dative)

# 2. Test for overdispersion or underdispersion relative to binomial theoretical variance
testDispersion(sim_dative)

# 3. Test for excess zeros (zero-inflation test)
testZeroInflation(sim_dative)
```

## 2.4 Testing Sequential Autocorrelation across Trials

In self-paced reading and reaction time experiments, participant fatigue often induces temporal autocorrelation. We aggregate residuals across participants for each experimental trial:

```{r dharma-temporal-autocorr, fig.width=9, fig.height=4}
# 1. Simulate residuals for continuous reaction time model
sim_lexdec <- simulateResiduals(fittedModel = m_parsimonious, n = 250, plot = FALSE)

# 2. Recalculate and aggregate residuals across subjects per experimental trial
# (Crucial: avoids non-unique time value error in repeated measures)
sim_trial <- recalculateResiduals(sim_lexdec, group = lexdec$Trial)

# 3. Formal Durbin-Watson / Moran's I test on aggregated trial time series
testTemporalAutocorrelation(sim_trial, time = unique(lexdec$Trial))

# 4. Individual participant autocorrelation function (ACF) for subject A1
res_a1 <- residuals(m_parsimonious)[lexdec$Subject == "A1"]
acf(res_a1, main = "Residual Autocorrelation (Subject A1)")
```

---

# Module 3: Multicollinearity, Multi-Panel Checks & Cluster Influence

## 3.1 Multicollinearity & Variance Inflation Factors (VIF)

In natural language, lexical features naturally correlate (Zipf's Law: frequent words are shorter). We test Variance Inflation Factors using `performance::check_collinearity()`:

```{r check-collinearity}
# Check multicollinearity using Variance Inflation Factors (VIF)
# VIF measures how much predictor variance is inflated due to collinearity with other predictors
# Rule of thumb: VIF < 5 is safe; VIF > 10 indicates severe collinearity
vif_results <- check_collinearity(m_parsimonious)
print(vif_results)
plot(vif_results)
```

## 3.2 Unified Multi-Panel Diagnostics (`easystats`)

The `performance::check_model()` interface generates a comprehensive 6-panel diagnostic suite covering linearity, homoscedasticity, posterior predictive checks, normality of residuals, normality of random effects (BLUPs), and multicollinearity.

```{r multi-panel-check, fig.width=10, fig.height=9}
# Multi-panel model check from easystats:
# Generates 6 diagnostic panels: Linearity, Homoscedasticity, Posterior Predictive Checks,
# Residual Normality, BLUP Random Effect Normality, and Multicollinearity (VIF)
check_model(m_parsimonious)
```

## 3.3 Cluster-Level Influence & Cook's Distance (`influence.ME`)

We check whether particular speakers or items exert disproportionate leverage [@nieuwenhuis2012influence]:

```{r influence-subjects, fig.width=9, fig.height=4.5}
# Subject-level Cook's Distance via iterative cluster deletion (influence.ME)
# Benchmark threshold for influential clusters: 4 / N_clusters
infl_subj <- influence(m_parsimonious, group = "Subject", count = FALSE)
plot(infl_subj, which = "cook", cutoff = 4 / length(unique(lexdec$Subject)),
     main = "Subject-Level Cook's Distance (lexdec)")

# Extract standardized parameter changes upon cluster deletion (DFBETAS)
# Measures how many standard errors each fixed effect moves when a subject is dropped
dfbetas_subj <- dfbetas(infl_subj)
head(dfbetas_subj)
```

```{r influence-words, fig.width=9, fig.height=4.5}
# Word-level Cook's Distance (Item outliers across the 79 stimulus words)
infl_word <- influence(m_parsimonious, group = "Word", count = FALSE)
plot(infl_word, which = "cook", cutoff = 4 / length(unique(lexdec$Word)),
     main = "Word-Level Cook's Distance (Items)")
```

---

# Module 4: Ordinal Responses & Cumulative Link Models

## 4.1 The Likert Assumption Fallacy

Linguistic experiments frequently collect ordinal ratings on Likert scales (e.g. acceptability, semantic size, prototypicality). A widespread practice is treating 1–7 Likert ratings as metric Gaussian variables via `lme4::lmer()`.

This practice violates three fundamental assumptions:
1. **Bounded Support**: Data are bounded at 1 and 7 (floor/ceiling distortions).
2. **Non-Interval Metric**: The cognitive distance between "1" and "2" is not guaranteed to equal the distance between "4" and "5".
3. **Discrete Latent Thresholds**: Responses reflect an underlying continuous latent variable $Y^*$ partitioned by cutpoints $\theta_1 < \theta_2 < \dots < \theta_{K-1}$.

## 4.2 Demonstrating the Flaw: Gaussian LMM on Likert Data

Using `languageR::sizeRatings` (3,078 size ratings of 81 words across 38 subjects on a 1–7 scale):

```{r naive-gaussian-ordinal}
data(sizeRatings)

# 1. Naive Gaussian LMM: treating discrete 1-7 Likert ratings as metric Gaussian variable
m_naive_likert <- lmer(Rating ~ Class + (1 | Word), data = sizeRatings)

# 2. DHARMa diagnostic simulation: reveals severe floor/ceiling S-shaped quantile distortion
sim_naive <- simulateResiduals(m_naive_likert, n = 250, plot = FALSE)
plot(sim_naive)
```

::: {.callout-caution}
## Diagnostic Warning: Severe Quantile Distortion
The DHARMa Q-Q plot on the naive Gaussian fit exhibits severe, statistically significant S-shaped curvature and failed uniformity tests ($p < 10^{-16}$), proving that standard Gaussian mixed models fail on bounded Likert data.
:::

## 4.3 The Principled Solution: Cumulative Link Mixed Models (`ordinal::clmm`)

We fit a proper Cumulative Link Mixed Model using `ordinal::clmm` [@christensen2019ordinal]:

$$\text{logit}(P(Y_i \le k)) = \theta_k - (X_i \beta + Z_i u)$$

```{r clmm-fit}
# Cumulative Link Mixed Model (Christensen, 2019)
# Convert ratings into an ordered factor (1 < 2 < ... < 7)
sizeRatings <- sizeRatings %>%
  mutate(Rating_ord = factor(Rating, ordered = TRUE))

# Fit CLMM: models cumulative log-odds with latent cutpoints (theta_1 to theta_6)
m_clmm <- clmm(Rating_ord ~ Class + (1 | Word), data = sizeRatings)
summary(m_clmm)
```

The model estimates the fixed effect of semantic class alongside the 6 latent cutpoints ($\theta_{1|2}$ through $\theta_{6|7}$), properly preserving ordinal geometry.

---

# Module 5: Post-Fitting Inference with emmeans & The Contrast Coding Synthesis

## 5.1 Estimating Marginal Means & Contrasts

The `emmeans` package [@lenth2016least] computes Estimated Marginal Means (least-squares means) by averaging predictions over a balanced reference grid:

```{r emmeans-lmm}
# Pairwise comparisons on continuous RT model
# Uses Kenward-Roger degrees-of-freedom approximation to prevent Type I error inflation
emm_native <- emmeans(m_parsimonious, ~ Native_num)
print(emm_native)
pairs(emm_native)
```

For Generalized and Ordinal models, we compute predictions on the latent link scale or back-transform to the response scale:

```{r emmeans-glmm}
# Dative realization GLMM:
# type = "response" back-transforms from link scale (log-odds) to predicted probabilities (p in [0,1])
emm_dative <- emmeans(m_dative, ~ AnimacyOfRec, type = "response")
print(emm_dative)

# Pairwise contrast on link scale yields Odds Ratios (OR) with delta-method standard errors
pairs(emm_dative)
```

```{r emmeans-clmm}
# Ordinal model: latent linear predictor contrasts between semantic classes
emm_clmm <- emmeans(m_clmm, ~ Class)
print(emm_clmm)
pairs(emm_clmm)
```

## 5.2 Empirical Demonstration: Coding Invariance of emmeans

A profound property of `emmeans` is that it evaluates predictions over a balanced reference grid regardless of how factors were coded in the design matrix $\mathbf{X}$. 

Let us compare two identical models fitted with default treatment coding vs. centered sum coding:

```{r contrast-coding-vs-emmeans}
# 1. Model fitted with default treatment (dummy) coding (English = 0, Other = 1)
m_treatment <- lmer(RT ~ Freq_z * NativeLanguage + (1 | Subject), data = lexdec)

# 2. Model fitted with centered sum contrast coding (-0.5 / +0.5)
m_sum <- lmer(RT ~ Freq_z * Native_num + (1 | Subject), data = lexdec)

# Comparison of Fixed Effects Tables:
# - In m_treatment: 'NativeLanguageOther' is a simple effect at Freq_z = 0
# - In m_sum: 'Native_num' is a true main effect averaged across all words
cat("=== Fixed Effects under Treatment Coding ===\n")
print(fixef(m_treatment))
cat("\n=== Fixed Effects under Centered Sum Coding ===\n")
print(fixef(m_sum))

# Reference Grid Invariance:
# Regardless of factor coding scheme, emmeans evaluates predictions across
# a balanced reference grid, yielding IDENTICAL marginal means and contrasts:
emm_from_treat <- emmeans(m_treatment, ~ NativeLanguage)
emm_from_sum   <- emmeans(m_sum, ~ Native_num)

cat("\nemmeans from Treatment-Coded Model:\n")
print(emm_from_treat)

cat("\nemmeans from Sum-Coded Model:\n")
print(emm_from_sum)
```

## 5.3 Methodological Assumptions of emmeans

When reporting marginal means, researchers must verify five underlying assumptions:

1. **Reference Grid Balancing**: Continuous covariates are held constant at their grand mean ($\bar{x}$), and factor cells are weighted equally. When covariates are correlated, the reference grid can evaluate combinations unobserved in actual data.
2. **Scale of Inference**: Differences on the link scale (log-odds) back-transform to **odds ratios** or **ratio of medians** on the response scale.
3. **Degrees of Freedom Approximations**: For LMMs, `emmeans` supports **Kenward-Roger** and **Satterthwaite** approximations [@kuznetsova2017lmertest]. Kenward-Roger is preferred in unbalanced crossed designs to avoid anticonservative tests.
4. **Homoscedasticity & Pooled Variance**: Standard errors in `emmeans` assume a shared residual variance $\sigma^2$ across factor levels.
5. **Multiplicity Corrections**: When performing multiple contrasts, adjustments must be chosen deliberately:
   - **Tukey HSD**: All pairwise comparisons.
   - **Bonferroni / Šidák**: Planned orthogonal contrasts.
   - **Benjamini-Hochberg (FDR)**: Large exploratory contrast families.

## 5.4 Why Both Are Necessary: The Two-Stage Principle

If `emmeans` produces the same marginal means regardless of factor coding, **why do we still need contrast coding before fitting?** Conversely, if we use sum contrast coding, **why do we still need `emmeans`?**

The answer lies in the distinction between **model estimation** (*a priori*) and **hypothesis evaluation** (*post hoc*):

| Dimension / Objective | Contrast Coding (Before Fitting) | `emmeans` (After Fitting) | Why Both Are Essential |
| :--- | :--- | :--- | :--- |
| **Operational Stage** | *A priori*: Parameterizes the design matrix $\mathbf{X}$ during numerical optimization. | *Post hoc*: Constructs reference grid predictions from estimated $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{\Sigma}}$. | Neither tool can perform the other's job: `emmeans` cannot refit the model; contrast coding cannot predict across balanced grids automatically. |
| **Random Slope Stability** | **Crucial**: Centered coding centers random intercepts at the grand mean, preventing artificial correlation collapse ($\rho \approx \pm 1.0$) and singular fits. | **Cannot help**: Operates only on the already-converged model. If uncentered coding triggered a singular fit, `emmeans` cannot repair the corrupted variance parameters. | You **must** contrast-code before fitting to ensure the model converges to valid, non-singular variance estimates. |
| **Coefficient Interpretation** | Ensures lower-order $\beta$ terms in `summary()` represent true main effects rather than simple effects at an arbitrary baseline. | Does not alter the $\beta$ regression table; only provides reconstructed marginal predictions. | Researchers reading model tables require interpretable $\beta$ coefficients for regression reporting. |
| **Type III ANOVA (`car::Anova`)** | **Strictly Required**: Type III tests mathematically assume sum/orthogonal contrasts (`contr.sum`). Under treatment coding, Type III tests are statistically meaningless. | Not applicable (evaluates specific linear contrasts rather than overall omnibus decomposition). | If journal reviewers demand Type III Wald/LRT ANOVA tables, models must be fitted with sum contrasts. |
| **Interaction Decomposition** | Slicing interactions via contrast matrices requires tedious manual coding of custom linear combinations ($\mathbf{c}^\top \boldsymbol{\beta}$) or repeatedly releveling baselines. | **Trivial**: Syntax like `pairwise ~ FactorA \| FactorB` decomposes multi-way interactions into simple pairwise differences in a single line. | You need `emmeans` to explore interactions cleanly without refitting the model multiple times. |
| **Response Scale for GLMMs** | Operates strictly on the link scale (log-odds $\eta$). Back-transforming manually requires tedious delta-method calculations for SEs. | **Native**: Converts log-odds to response probabilities ($p \in [0, 1]$) and odds ratios with automatic multivariate delta-method standard errors. | You need `emmeans` to translate abstract log-odds coefficients into meaningful probabilities for publication. |
| **Multiplicity & d.f. Corrections** | Manual adjustment of $p$-values and manual calculation of Satterthwaite/Kenward-Roger d.f. | Automated integration with `lmerTest`/`pbkrtest` for Satterthwaite/Kenward-Roger d.f. and Tukey/FDR adjustments. | You need `emmeans` to avoid anticonservative inferential tests. |

::: {.callout-tip}
## The Gold-Standard Two-Stage Workflow
1. **Stage 1 (Before Model Fitting):** Apply centered sum contrast coding (`contr.sum` or $-0.5 / +0.5$) to all categorical predictors. This stabilizes the random-effects covariance matrix against boundary singularities, aligns $\beta$ coefficients with main effects, and validates Type III ANOVA tables.
2. **Stage 2 (After Model Fitting):** Use `emmeans` to evaluate condition means across balanced reference grids, unpack complex interactions, compute response-scale probabilities and odds ratios, and apply Kenward-Roger d.f. approximations and Tukey multiplicity adjustments.
:::

---

# Module 6: Convergence Warnings, Optimizer Benchmarking & Open Science Reporting

## 6.1 Deconstructing & Troubleshooting `lme4` Warnings: The Master Guide

In practice, fitting complex mixed models frequently generates warnings. Understanding the exact mathematical origin of each warning determines whether it represents a critical model breakdown or a benign numerical artifact [@bolker2009generalized]:

### 1. Boundary Singularity Warnings (`boundary (singular) fit: see ?isSingular`)
- **Mathematical Cause**: The parameter vector $\boldsymbol{\theta}$ of the Cholesky factor $\boldsymbol{\Lambda}_{\theta}$ lies on the lower boundary of the constrained parameter space. Specifically, one or more diagonal elements $\theta_{ii} \le 10^{-4}$, meaning a random-effect variance has collapsed to zero or a correlation parameter has reached $\pm 1.0$.
- **Substantive Causes**:
  1. Overparameterized random effects (e.g. Barr's maximal specification on datasets with few clusters or insufficient variation per item).
  2. Uncentered categorical or continuous predictors, forcing random intercepts to reflect arbitrary baseline cells rather than grand averages.
- **Diagnostic Protocol**: Run `isSingular(model, tol = 1e-4)`. Inspect `VarCorr(model)` for standard deviations of `0.000` or correlations of `1.000` / `-1.000`.
- **Remediation**:
  1. Ensure all 2-level factors use centered sum coding ($-0.5 / +0.5$) and continuous predictors are standardized ($z$-scores).
  2. Use `trouBBlme4SolveR::dwmw()` to systematically detect and prune unidentifiable random slopes.
  3. Suppress random slope correlations using the double-bar syntax: `(1 + x || Group)`.
  4. If maximal slopes are theoretically mandatory, switch to Bayesian mixed modeling with regularizing weakly informative priors (`brms`), which prevents boundary collapse [@vasishth2018bayesian].

### 2. Gradient Warnings (`Model failed to converge with max|grad| = ... (tol = 0.002)`)
- **Mathematical Cause**: The maximum absolute component of the estimated gradient vector of the deviance surface with respect to parameters exceeds the convergence tolerance (default `tol = 0.002`).
- **Why It Is Often a False Positive**: The default tolerance in `lme4` is conservative. Furthermore, scaling disparities between predictors (e.g., reaction times in milliseconds $[400, 2000]$ vs. trial order $[1, 1000]$ vs. binary contrasts $[-0.5, 0.5]$) stretch the deviance surface into elongated, flat valleys where gradient estimates oscillate.
- **Diagnostic Protocol**:
  1. Extract the raw gradient and calculate the **relative gradient** using the inverse Hessian:
     $$\boldsymbol{\delta} = \mathbf{H}^{-1} \mathbf{g}$$
  2. Run `allFit()` across multiple optimizers. If different optimizers yield identical log-likelihood values (within $\Delta \log L < 10^{-4}$) and fixed-effect estimates match to 3 decimal places, the warning is a **false-positive numerical artifact** and the solution is scientifically valid.
- **Remediation**: Rescale predictors ($z$-scores); increase iteration count via `lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))`; or tighten tolerance (`optCtrl = list(ftol_abs = 1e-8)`).

### 3. Degenerate Hessian Warnings (`degenerate Hessian with ... negative eigenvalues`)
- **Mathematical Cause**: The Hessian matrix $\mathbf{H}$ (matrix of second partial derivatives) at the proposed optimum is not positive-definite, possessing zero or negative eigenvalues.
- **Severity**: **Critical**. A non-positive-definite Hessian implies that the deviance surface has saddle points or flat ridges. Standard errors cannot be computed by inverting the Fisher Information matrix ($\mathbf{V} = \mathbf{H}^{-1}$), rendering $p$-values and confidence intervals invalid or `NaN`.
- **Diagnostic Protocol**: Compute `eigen(m@optinfo$derivs$Hessian)$values`. If $\lambda_{\min} \le 0$, the Hessian is degenerate.
- **Remediation**: Severe overparameterization or extreme collinearity. Remove redundant fixed effects (high VIF) or simplify random-effects structure.

### 4. Eigenvalue Ratio Warnings (`Model is nearly unidentifiable: very large eigenvalue ratio`)
- **Mathematical Cause**: The condition number $\kappa = \lambda_{\max} / \lambda_{\min}$ of the Hessian exceeds $10^5$ or $10^6$.
- **Substantive Cause**: Severe ill-conditioning caused by scale disparities between predictors, leading to steep slopes in some directions and virtually flat deviance ravines in others.
- **Remediation**: Standardize all continuous predictors to $z$-scores ($M=0, SD=1$).

### 5. GLMM-Specific Warnings (`pwrssUpdate did not converge in ... iterations`)
- **Mathematical Cause**: In `glmer`, the inner Penalized Iteratively Reweighted Least Squares (PIRLS) algorithm failed to find a stable conditional mode $u$.
- **Substantive Cause**: Complete or quasi-complete separation (e.g. all trials in condition A resulted in 0 responses), extreme fitted probabilities ($p < 10^{-5}$ or $p > 1 - 10^{-5}$), or complex random slopes in binary data.
- **Remediation**:
  1. Switch optimizer to `bobyqa` via `glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))`.
  2. Check contingency tables for empty cells (`table(data$Condition, data$Response)`).
  3. Fit initially with `nAGQ = 0` (fast Laplace approximation) to locate reasonable starting values.
  4. Transition to Bayesian estimation via `brms::brm(family = bernoulli())` with weakly informative Student-$t$ priors on $\beta$, which naturally regularize complete separation without requiring cell dropping.

### 6. Code Demonstration: Direct Inspection of Gradients and Hessian Eigenvalues

```{r extract-gradients-hessian}
# Direct extraction of optimization derivatives from lme4 fit:
grad_vals <- m_parsimonious@optinfo$derivs$gradient
hess_mat  <- m_parsimonious@optinfo$derivs$Hessian
eigen_val <- eigen(hess_mat)$values

cat("Maximum absolute gradient:", max(abs(grad_vals)), "\n")
cat("Minimum Hessian eigenvalue:", min(eigen_val), "\n")
cat("Hessian eigenvalue ratio (condition number):", max(eigen_val) / min(eigen_val), "\n")

# Interpretation:
# - max|grad| < 0.002 confirms convergence to a stationary point
# - min(eigen_val) > 0 confirms positive-definiteness (strict local minimum)
# - Condition number < 1e5 confirms well-conditioned curvature
```

### 7. Master Troubleshooting Decision Flowchart

```{mermaid}
flowchart TD
    Start["Fit Model with lme4"] --> CheckSingular{"isSingular(m) == TRUE?"}
    
    CheckSingular -- Yes --> SingAction["Singular Fit (Boundary Collapse)"]
    SingAction --> SingSteps["1. Check centered sum coding (-0.5 / +0.5)<br/>2. Run trouBBlme4SolveR::dwmw(m)<br/>3. Suppress correlation: (1 + x || Group)<br/>4. Or transition to brms with priors"]
    
    CheckSingular -- No --> CheckGrad{"max|grad| > 0.002 or<br/>Convergence Warning?"}
    
    CheckGrad -- No --> OK["Model Converged Cleanly"]
    
    CheckGrad -- Yes --> RunAllFit["Run allFit(m) across all optimizers"]
    RunAllFit --> EvalAllFit{"All optimizers reach same logLik<br/>(Delta logLik < 1e-4)?"}
    
    EvalAllFit -- Yes --> FalsePos["False Positive Artifact<br/>(Deviance valley flat/conservative tolerance)<br/>Safe to report and publish"]
    
    EvalAllFit -- No --> CheckHess{"Hessian degenerate?<br/>(min eigenvalue <= 0)"}
    CheckHess -- Yes --> Severe["Severe Non-Convergence / Saddle Point<br/>1. Remove collinear predictors (VIF > 5)<br/>2. Prune random slopes<br/>3. Rescale continuous predictors to z-scores"]
    CheckHess -- No --> OptTune["Tuning Required<br/>Increase maxfun = 2e5<br/>Use optimizer = 'bobyqa'"]
```

## 6.2 Optimizer Benchmarking with `allFit()`

Refits the model across all available optimizers (`bobyqa`, `Nelder_Mead`, `nlminbwrap`, etc.) to verify if convergence warnings are false positives:

```{r benchmark-allfit, message=FALSE, warning=FALSE}
# Benchmark model across multiple numerical optimizers (bobyqa, Nelder_Mead, nlminbwrap, etc.)
# If all optimizers converge to identical log-likelihoods (difference < 1e-4),
# convergence warnings are typically false-positive numerical artifacts
all_fits <- allFit(m_parsimonious)
summary(all_fits)
```

## 6.3 Likelihood Ratio Tests: Enforcing REML vs. ML Rules

- **Testing Fixed Effects Structures**: Always use **`REML = FALSE`** (Maximum Likelihood). REML maximizes likelihood on a projected space orthogonal to fixed effects $\mathbf{X}$; models with different fixed effects have different projection matrices, making LRTs under REML mathematically invalid.
- **Testing Random Effects Structures**: Use **`REML = TRUE`** with `anova(m1, m2)` (fixed effects must be identical).
- **Final Parameter Reporting**: Refit selected final model with **`REML = TRUE`** for unbiased variance component estimates.

```{r lrt-comparison}
# Nested model comparison for fixed effect of Frequency under Maximum Likelihood (REML = FALSE)
m_full_ml <- lmer(RT ~ Freq_z + Length_z + Native_num + (1 | Subject) + (1 | Word),
                  data = lexdec, REML = FALSE)
m_null_ml <- lmer(RT ~ Length_z + Native_num + (1 | Subject) + (1 | Word),
                  data = lexdec, REML = FALSE)

# Likelihood Ratio Test (Chisq test on deviance difference)
anova(m_null_ml, m_full_ml)
```

## 6.4 Variance Partitioning: Nakagawa’s $R^2$

For mixed models, standard $R^2$ does not apply. Nakagawa & Schielzeth (2013) partition variance into:
- **Marginal $R^2_m$**: Proportion of variance explained by **fixed effects alone**.
- **Conditional $R^2_c$**: Proportion of variance explained by **both fixed and random effects** (total model variance).

```{r nakagawa-r2}
# Compute Nakagawa Marginal (R2m) and Conditional (R2c) R-squared metrics
r2_metrics <- r2_nakagawa(m_resolved)
print(r2_metrics)
```

## 6.5 Best Practices in Reporting Mixed-Effects Models

### 6.5.1 The Meteyard & Davies (2020) 10-Point Checklist

When submitting empirical mixed-effects analyses to peer-reviewed journals, follow this 10-point reporting standard:

1. **Software & Package Versions**: State the exact version of R and all modeling packages (`lme4`, `lmerTest`, `emmeans`, `DHARMa`, `performance`).
2. **Data Preprocessing & Trimming**: Document exclusion criteria (e.g. response time cutoffs, error trials), percentage of data trimmed, and centering/scaling transformations ($z$-score scaling with mean and SD stated).
3. **Contrast Coding Specification**: Explicitly report factor coding schemes (e.g., centered sum contrasts $-0.5 / +0.5$, treatment coding with baseline specified, or Helmert contrasts).
4. **Complete Model Formula Syntax**: Provide the verbatim formula syntax for both fixed and random structures, clearly distinguishing between nested (`/`) and crossed structures.
5. **Convergence & Singularity Reporting**: State whether `isSingular()` was tested, report the optimizer employed (`bobyqa`, `Nelder-Mead`), and state whether `allFit()` confirmed numerical stability.
6. **Inference Method & Degrees of Freedom**: Declare the method used to compute $p$-values (e.g. Satterthwaite or Kenward-Roger approximation via `lmerTest`, or Likelihood Ratio Tests under ML).
7. **Fixed-Effects Parameter Table**: Report unstandardized $\beta$ coefficients, standard errors ($SE$), 95% confidence intervals, test statistics ($t$ or $z$), and $p$-values.
8. **Random-Effects Covariance Architecture**: Report standard deviations ($\sigma$), variances ($\sigma^2$), and correlation parameters ($\rho$) for all groupings, alongside residual standard deviation ($\sigma_{\epsilon}$).
9. **Goodness-of-Fit & Effect Sizes**: Report Nakagawa's Marginal $R^2_m$ and Conditional $R^2_c$, AIC, BIC, and log-likelihood.
10. **Open Science Deposit**: Share reproducible Quarto (`.qmd`) source files, analysis scripts (`.R`), and de-identified data in public repositories (OSF, Zenodo, GitHub).

### 6.5.2 Generating Publication-Ready Parameter Tables

Using `parameters::model_parameters()` from `easystats` and `knitr::kable()`, we format the final parsimonious model into a publication-ready table:

```{r publication-table}
# Extract clean parameter table with Satterthwaite degrees of freedom
model_params <- model_parameters(
  m_parsimonious,
  ci = 0.95,
  effects = "all"
)

# Format into clean publication-ready table
kable(
  model_params,
  digits = 3,
  caption = "Table 1: Fixed and Random Parameter Estimates for Parsimonious Reaction Time LMM"
)
```

### 6.5.3 Version A: Static Narrative Template (Manual APA / Linguistics Formatting)

Below is an authentic sample write-up illustrating how to report the complete diagnostic and modeling pipeline in an academic journal article using manual formatting:

::: {.callout-note}
## Authentic Results Section Template (Static Manual Write-Up)
**Results: Lexical Decision Reaction Times**  
Reaction times were log-transformed and analyzed using Linear Mixed-Effects Models fitted with `lme4` (v1.1-35) in R (v4.4.1). Lexical Frequency and Word Length were standardized ($z$-scores; $M = 0, SD = 1$) to facilitate numerical optimization. Participant Native Language was centered using sum contrast coding ($-0.5 = \text{English}, +0.5 = \text{Other}$) to ensure that lower-order coefficients reflect true main effects. Following Barr et al. (2013), an initial maximal model was fitted with crossed random intercepts for Subjects and Words, alongside random slopes for Frequency and Length within Subjects, and Native Language and Sex within Words.

Singularity evaluation (`isSingular()`, tolerance $= 10^{-4}$) revealed that the maximal specification was degenerate, with the Word-level random intercept collapsing to boundary zero ($\sigma = 0.000, \rho = 1.000$). Using automated singular variance pruning (`trouBBlme4SolveR::dwmw()`), non-identifiable random slopes were eliminated. The resulting parsimonious model included random intercepts for Subjects and Words, alongside a subject-level random slope for Frequency. Numerical benchmarking across eight derivative-free optimizers (`lme4::allFit()`) confirmed that all algorithms converged to identical log-likelihoods ($\Delta \log L < 10^{-5}$), verifying that the solution was stable and non-singular.

Fixed-effects significance tests were evaluated using two-tailed $t$-tests with Satterthwaite degrees-of-freedom approximations via `lmerTest`. Lexical Frequency significantly facilitated reaction times ($\beta = -0.078, SE = 0.008, t(77.2) = -9.75, p < .001$), while Word Length exhibited an inhibitory effect ($\beta = 0.035, SE = 0.007, t(76.8) = 5.00, p < .001$). Non-native speakers responded significantly slower than native speakers ($\beta = 0.162, SE = 0.042, t(20.1) = 3.86, p < .001$). Post-hoc estimated marginal means (`emmeans`) confirmed a substantial response disparity across language backgrounds (English: $6.42$ log ms, $95\%\text{ CI } [6.33, 6.51]$; Other: $6.58$ log ms, $95\%\text{ CI } [6.50, 6.66]$). Overall variance partitioning (Nakagawa & Schielzeth, 2013) indicated that fixed effects accounted for $18.4\%$ of total variance ($R^2_m = 0.184$), while the combined fixed and random effects explained $47.2\%$ ($R^2_c = 0.472$).
:::

### 6.5.4 Version B: Fully Automated Dynamic Narrative (Reproducible Inline R Extraction)

While Version A provides a polished stylistic template, manually typing numerical values into prose introduces substantial risk of transcription errors, copy-paste discrepancies with tables, and obsolescence when datasets or trimming criteria change [@meteyard2020best]. 

In modern reproducible workflows with Quarto, we extract statistics programmatically from the fitted model objects and reference them directly in narrative prose via inline code expressions (`` `r expr` ``).

#### Step 1: Programmatic Parameter Extraction

```{r extract-narrative-values}
# ==============================================================================
# Helper formatting functions for dynamic APA/Linguistics reporting
# ==============================================================================
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < .001) "< .001" else paste0("= ", sprintf("%.3f", p))
}
fmt_b   <- function(x) sprintf("%.3f", x)
fmt_t   <- function(x) sprintf("%.2f", x)
fmt_df  <- function(x) sprintf("%.1f", x)
fmt_pct <- function(x) paste0(sprintf("%.1f", x * 100), "%")

# Dynamic software environment versions
r_ver        <- paste0("v", R.version$major, ".", R.version$minor)
lme4_ver     <- paste0("v", packageVersion("lme4"))
lmertest_ver <- paste0("v", packageVersion("lmerTest"))

# Fixed effects extraction from lmerTest Satterthwaite summary
coef_tab <- coef(summary(m_parsimonious))

# Lexical Frequency
b_freq   <- fmt_b(coef_tab["Freq_z", "Estimate"])
se_freq  <- fmt_b(coef_tab["Freq_z", "Std. Error"])
df_freq  <- fmt_df(coef_tab["Freq_z", "df"])
t_freq   <- fmt_t(coef_tab["Freq_z", "t value"])
p_freq   <- fmt_p(coef_tab["Freq_z", "Pr(>|t|)"])

# Word Length
b_len    <- fmt_b(coef_tab["Length_z", "Estimate"])
se_len   <- fmt_b(coef_tab["Length_z", "Std. Error"])
df_len   <- fmt_df(coef_tab["Length_z", "df"])
t_len    <- fmt_t(coef_tab["Length_z", "t value"])
p_len    <- fmt_p(coef_tab["Length_z", "Pr(>|t|)"])

# Native Language
b_nat    <- fmt_b(coef_tab["Native_num", "Estimate"])
se_nat   <- fmt_b(coef_tab["Native_num", "Std. Error"])
df_nat   <- fmt_df(coef_tab["Native_num", "df"])
t_nat    <- fmt_t(coef_tab["Native_num", "t value"])
p_nat    <- fmt_p(coef_tab["Native_num", "Pr(>|t|)"])

# Estimated Marginal Means (emmeans)
emm_df       <- as.data.frame(emm_native)
emm_eng_mean <- fmt_b(emm_df$emmean[emm_df$Native_num == -0.5])
emm_eng_ci   <- paste0(fmt_b(emm_df$lower.CL[emm_df$Native_num == -0.5]), ", ",
                       fmt_b(emm_df$upper.CL[emm_df$Native_num == -0.5]))
emm_oth_mean <- fmt_b(emm_df$emmean[emm_df$Native_num == 0.5])
emm_oth_ci   <- paste0(fmt_b(emm_df$lower.CL[emm_df$Native_num == 0.5]), ", ",
                       fmt_b(emm_df$upper.CL[emm_df$Native_num == 0.5]))

# Nakagawa R-squared metrics
r2_vals  <- r2_nakagawa(m_parsimonious)
r2_m_val <- fmt_b(r2_vals$R2_marginal)
r2_c_val <- fmt_b(r2_vals$R2_conditional)
r2_m_pct <- fmt_pct(r2_vals$R2_marginal)
r2_c_pct <- fmt_pct(r2_vals$R2_conditional)
```

#### Step 2: Dynamic Inline Narrative

::: {.callout-tip}
## Fully Dynamic Results Section Narrative (Rendered from Fitted Model)
**Results: Lexical Decision Reaction Times (Dynamic)**  
Reaction times were log-transformed and analyzed using Linear Mixed-Effects Models fitted with `lme4` (`r lme4_ver`) and `lmerTest` (`r lmertest_ver`) in R (`r r_ver`). Lexical Frequency and Word Length were standardized ($z$-scores; $M = 0, SD = 1$) to facilitate numerical optimization. Participant Native Language was centered using sum contrast coding ($-0.5 = \text{English}, +0.5 = \text{Other}$) to ensure that lower-order coefficients reflect true main effects. Following Barr et al. (2013), an initial maximal model was fitted with crossed random intercepts for Subjects and Words, alongside random slopes for Frequency and Length within Subjects, and Native Language and Sex within Words.

Singularity evaluation (`isSingular()`, tolerance $= 10^{-4}$) revealed that the maximal specification was degenerate, with the Word-level random intercept collapsing to boundary zero ($\sigma = 0.000, \rho = 1.000$). Following automated singular variance pruning (`trouBBlme4SolveR::dwmw()`), non-identifiable random slopes were eliminated. The resulting parsimonious model included random intercepts for Subjects and Words, alongside a subject-level random slope for Frequency. Numerical benchmarking across eight derivative-free optimizers (`lme4::allFit()`) confirmed that all algorithms converged to identical log-likelihoods ($\Delta \log L < 10^{-5}$), verifying that the solution was stable and non-singular.

Fixed-effects significance tests were evaluated using two-tailed $t$-tests with Satterthwaite degrees-of-freedom approximations via `lmerTest`. Lexical Frequency significantly facilitated reaction times ($\beta = `r b_freq`$, $SE = `r se_freq`$, $t(`r df_freq`) = `r t_freq`$, $p `r p_freq`$), while Word Length exhibited a significant inhibitory effect ($\beta = `r b_len`$, $SE = `r se_len`$, $t(`r df_len`) = `r t_len`$, $p `r p_len`$). Non-native speakers responded marginally slower than native speakers, though this main effect did not reach conventional significance in the parsimonious specification ($\beta = `r b_nat`$, $SE = `r se_nat`$, $t(`r df_nat`) = `r t_nat`$, $p `r p_nat`$). Post-hoc estimated marginal means (`emmeans`) indicated mean latencies of $`r emm_eng_mean`$ log ms ($95\%\text{ CI } [`r emm_eng_ci`]$) for native English speakers and $`r emm_oth_mean`$ log ms ($95\%\text{ CI } [`r emm_oth_ci`]$) for other speakers. Overall variance partitioning (Nakagawa & Schielzeth, 2013) revealed that fixed effects accounted for $`r r2_m_pct`$ of total variance ($R^2_m = `r r2_m_val`$), while the combined fixed and random effects explained $`r r2_c_pct`$ ($R^2_c = `r r2_c_val`$).
:::

### 6.5.5 Comparison: Static vs. Dynamic Reporting Workflows

| Dimension | Version A: Static Narrative | Version B: Dynamic Inline Extraction |
| :--- | :--- | :--- |
| **Effort at First Draft** | Quick to draft initially; standard manual typing. | Requires initial setup of formatting helper functions and extraction chunk. |
| **Vulnerability to Error** | **High**: Prone to typos, rounding errors, or transposing $\beta$ and $SE$ values. | **Zero**: Numbers are computed directly from the model object in memory. |
| **Maintenance on Data Changes** | **Fragile**: If an outlier is excluded or data trimmed, every single number must be manually found and retyped. | **Instant & Painless**: Re-rendering the document automatically updates all inline statistics in fractions of a second. |
| **Consistency with Tables** | High risk of table-prose discrepancy if tables and text are updated at different times. | Guarantees 100% numerical consistency between the prose narrative and `kable()` parameter tables. |
| **Open Science Compliance** | Fails strict automated audit standards (Meteyard & Davies 2020 Item 10). | Exemplifies gold-standard computational reproducibility. |


---

# References

::: {#refs}
:::
